---
title: "Final project, Beta diversity. "
output: html_document
---

# Step 1)

Load libraries and Data, I rarefied the data based on the conclusions we came to when looking at the QUIME results.The rarefaction number is 5396

```{r Load data and libraries}
library(tidyverse)
library(phyloseq)
library(vegan)
jg;ihjg;hg
# Load object
ps <- readRDS('data/phyloseq_taxonomy.rds')

# rarefy the data:
psrare <- ps %>%
          rarefy_even_depth(sample.size = 5396, rngseed = 421)
table(sample_sums(psrare))


```

#Step 2) Caclculate the beta diversity using the Bray-Curtis distance matrix.

```{r calculate bray-curtis distances}
# Calculate the distance between each sample pair
ps_WUF <- phyloseq::distance(psrare, method = "wunifrac")
View(as.matrix(ps_WUF)) # Every pairwise comparison included
```

#Step 3) use Multidimensional scaling to reduce the distance matrix into fewer dimentions which will help plot abd understand the Data.

```{r normalize and plot}
# MDS scaling
set.seed(421)
mds = metaMDS(ps_WUF)

# Extract data
mds_data = mds$points %>% as.data.frame %>%
  merge(sample_data(psrare), by='row.names', sort=F)
head(mds_data)
```

#Step 4) Plot the data based on the following comparisons, different HA/NA subtypes, different HA subtypes, different NA subtypes, finally Infected Vs non infection comparison.

```{r}
p_HA_NA = mds_data %>%
  ggplot(aes(MDS1,MDS2,color = HA_NA)) +
  geom_point() +
  stat_ellipse() +
  theme_classic(base_size=18)

p_HA = mds_data %>%
  ggplot(aes(MDS1,MDS2,color = ha_subtype)) +
  geom_point() +
  stat_ellipse() +
  theme_classic(base_size=18)


p_NA = mds_data %>%
  ggplot(aes(MDS1,MDS2,color = na_subtype)) +
  geom_point() +
  stat_ellipse() +
  theme_classic(base_size=18)

p_IN = mds_data %>%
  ggplot(aes(MDS1,MDS2,color = influenza_status)) +
  geom_point() +
  stat_ellipse() +
  theme_classic(base_size=18)

p_HA_NA

p_HA

p_NA

p_IN

```

#Step 5)

Perform a PERMANOVA test to find statistical significance between the groups using an Adonis function.

```{r}
metadata = sample_data(psrare) %>% data.frame() # to make the next lines easier

set.seed(420)
# Single variable
# when I do this with 999 permutations HA subtype become significant at p=0.05 but when I do it at 99999 permutations it is not significant p=0.054. IDK what to do 

ad_HA_NA <- adonis2(ps_WUF ~ HA_NA, data = metadata)

ad_HA <- adonis2(ps_WUF ~ ha_subtype, data = metadata)

ad_NA <- adonis2(ps_WUF ~ na_subtype, data = metadata)

ad_inf <- adonis2(ps_WUF ~ influenza_status, data = metadata)



ad_HA_NA99 <- adonis2(ps_WUF ~ HA_NA, data = metadata, permutations = 99999)

ad_HA99 <- adonis2(ps_WUF ~ ha_subtype, data = metadata, permutations = 99999)

ad_NA99 <- adonis2(ps_WUF ~ na_subtype, data = metadata, permutations = 99999)

ad_inf99 <- adonis2(ps_WUF ~ influenza_status, data = metadata, permutations = 99999)


ad_HA_NA
ad_HA
ad_NA
ad_inf

ad_HA_NA99
ad_HA99
ad_NA99
ad_inf99

```

#removing the Neg form the sample data

```{r attempt 1 - subset metadata}


# When I remove the negative (uninfected) data points it looks nicer.
# This also makes it the HA subtype more significant. 

# MDS scaling
set.seed(420)
mds <- metaMDS(ps_WUF)

# Extract data
mds_data <- mds$points %>% as.data.frame %>%
  merge(sample_data(psrare), by='row.names', sort=F) 
mds_data_f <- filter( mds_data, influenza_status == "Pos")
  
head(mds_data_f)

# Plot it
p_HA_NA_F = mds_data_f %>%
  ggplot(aes(MDS1,MDS2,color = HA_NA)) +
  geom_point() +
  stat_ellipse() +
  theme_classic(base_size=18)

p_HA_F = mds_data_f %>%
  ggplot(aes(MDS1,MDS2,color = ha_subtype)) +
  geom_point() +
  stat_ellipse() +
  theme_classic(base_size=18)


p_NA_F = mds_data_f %>%
  ggplot(aes(MDS1,MDS2,color = na_subtype)) +
  geom_point() +
  stat_ellipse() +
  theme_classic(base_size=18)


p_HA_NA_F

p_HA_F

p_NA_F

# # IDK how to do this part 
# metadata_F<-metadata %>% filter( influenza_status == "Pos")
# ad_HA_NA_f <- adonis2(ps_WUF ~ HA_NA, data = mds_data_f)
# 
# ad_HA_f <- adonis2(ps_WUF ~ ha_subtype, data = mds_data_f)
# 
# ad_HA_f2<- adonis2(ps_WUF ~ ha_subtype, data = mds_data_f, permutations = 99999)
# 
# ad_NA_f <- adonis2(ps_WUF ~ na_subtype, data = mds_data_f)
# ad_HA_NA_f
# ad_HA_f
# ad_NA_f

```

```{r save outputs}
# # Save plot
# ggsave('../Results/Plots/Module 13c - Beta Diversity.jpeg',
#        plot = p,
#        height= 4, width = 6)
# 
# # Save statistics as an Excel file
# writexl::write_xlsx(list('Beta Diversity' = stats),
#                     '../Results/Tables/Module 13c - Beta Diversity.xlsx')
```
