---
title: "Alpha_Diversity"
output: html_document
---

### Introduction ###

The alpha diversity of cloacal microbiome data from influenza A–infected mallard ducks was analyzed using the Shannon and Faith’s diversity metrics. The code
for this analysis was adapted from UBC MICB 305 and MICB 205 course material.

Loading the packages needed for Alpha diversity measures and statistical 
analysis.

```{r}
library(tidyverse)
library(phyloseq)
library(vegan)
library(picante)
library(ape)
library(dunn.test)
library(ggsignif)
```

Importing processed phyloseq object from generate_phyloseq.Rmd
```{r}

ps <- readRDS('../processed_data/phyloseq_taxonomy.rds')

```
Rarefying the data.
```{r}
psrare= ps %>% rarefy_even_depth(sample.size = 5396, rngseed = 360)

#The rarefraction depth was decided based on qiime 2 outputs 
``` 
### Shannon's Diversity Index ###

## HA and NA ##
Calculating the Shannon diversity metric.
```{r}
set.seed(360)
p = plot_richness(psrare, x="HA_NA",
                        measures = c("Shannon"),
                  color= "HA_NA")

pdata = p$data
#Extracting Shannon diversity 

```
Statistical analysis of the shannon diversity metric between subtypes. 
```{r}
df <- data.frame(HA_NA = as.factor(pdata$HA_NA), value = pdata$value)
#Kruskal-Wallis and Dunn's test as post-hoc
res_sig <-dunn.test(df$value, df$HA_NA, kw=TRUE, method = "bh")

results_df <- data.frame(
  subtype_comparison = res_sig$comparisons,
  p_adjusted = res_sig$P.adjusted
)
```
## NA Not Aggregated ##
Shannon diversity calculation but grouped by NA subtype. 
```{r}
set.seed(360)
p_NA_shannon = plot_richness(psrare, x="na_subtype",
                        measures = c("Shannon"),
                  color= "na_subtype")

pdata_NA_shannon = p_NA_shannon$data

```
Statistical analysis of the shannon diversity metric between NA subtypes.
```{r}
df_NA_shannon <- data.frame(na_subtype = as.factor(pdata_NA_shannon$na_subtype), value = pdata_NA_shannon$value)
#Kruskal-Wallis and Dunn's test as post-hoc
res_sig_NA <-dunn.test(df_NA_shannon$value, df_NA_shannon$na_subtype, kw=TRUE,
                    method = "bh")

results_df_NA <- data.frame(
  subtype_comparison = res_sig_NA$comparisons,
  p_adjusted = res_sig_NA$P.adjusted
)
```
## HA Not Aggregated ##
Shannon diversity calculation but grouped by HA subtype. 
```{r}
set.seed(360)
p_HA_shannon = plot_richness(psrare, x="ha_subtype",
                        measures = c("Shannon"),
                  color= "ha_subtype")

pdata_HA_shannon = p_HA_shannon$data
```
Statistical analysis of the shannon diversity metric between HA subtypes.
```{r}
df_HA_shannon <- data.frame(ha_subtype = as.factor(pdata_HA_shannon$ha_subtype), value = pdata_HA_shannon$value)
#Kruskal-Wallis and Dunn's test as post-hoc
res_sig_HA <-dunn.test(df_HA_shannon$value, df_HA_shannon$ha_subtype, kw=TRUE,
                    method = "bh")

results_df_HA <- data.frame(
  subtype_comparison = res_sig_HA$comparisons,
  p_adjusted = res_sig_HA$P.adjusted
)
```
## NA Subtype Aggregated ##
Shannon diversity calculation but grouped by NA subtype with aggregated 
sub-subtype----->subtype. 
```{r}
set.seed(360)
p_NA_aggr_shannon = plot_richness(psrare, x="na_aggr",
                        measures = c("Shannon"),
                  color= "na_aggr")

pdata_NA_aggr_shannon = p_NA_aggr_shannon$data

p_formatted_NA_aggr = pdata_NA_aggr_shannon %>% ggplot(aes(na_aggr,value, fill = na_aggr)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(height= 0, width = 0.20, size = 1.5, alpha= 0.4) +
theme_classic(base_size=16)+
  ylab("Shannon Diversity") + xlab(NULL) +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1, size=12)) +
  labs(fill= "na_aggr")
p_formatted_NA_aggr

```
Statistical analysis of the shannon diversity metric between NA aggregated 
subtypes
```{r}
df_NA_aggr_shannon <- data.frame(na_aggr = as.factor(pdata_NA_aggr_shannon$na_aggr), value = pdata_NA_aggr_shannon$value)
#Kruskal-Wallis and Dunn's test as post-hoc
res_sig_aggr_NA <-dunn.test(df_NA_aggr_shannon$value, df_NA_aggr_shannon$na_aggr, kw=TRUE, method = "bh")
#one significant group
```
## HA_subtype Aggregrated ##
Shannon diversity calculation but grouped by HA subtype with aggregated 
sub-subtype----->subtype. 
```{r}
set.seed(360)
p_HA_aggr_shannon = plot_richness(psrare, x="ha_aggr",
                        measures = c("Shannon"),
                  color= "ha_aggr")

pdata_HA_aggr_shannon = p_HA_aggr_shannon$data

p_formatted_HA_aggr = pdata_HA_aggr_shannon %>% ggplot(aes(ha_aggr,value, fill = ha_aggr)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(height= 0, width = 0.20, size = 1.5, alpha= 0.4) +
theme_classic(base_size=16)+
  ylab("Shannon Diversity") + xlab(NULL) +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1, size=12)) +
  labs(fill= "ha_aggr")
p_formatted_HA_aggr

```
Statistical analysis of the shannon diversity metric between HA aggregated 
subtypes
```{r}
df_HA_aggr_shannon <- data.frame(ha_aggr = as.factor(pdata_HA_aggr_shannon$ha_aggr), value = pdata_HA_aggr_shannon$value)
#Kruskal-Wallis and Dunn's test as post-hoc
res_sig_aggr_HA <-dunn.test(df_HA_aggr_shannon$value, df_HA_aggr_shannon$ha_aggr, kw=TRUE, method = "bh")

#one significant group

```
## HA Phylogenetic Groups ##
Shannon diversity calculation but grouped by HA phylogenetic groups. 
```{r}
set.seed(360)
p_HA_group_shannon = plot_richness(psrare, x="ha_group",
                        measures = c("Shannon"),
                  color= "ha_group")

pdata_HA_group_shannon = p_HA_group_shannon$data

p_formatted_HA_group = pdata_HA_group_shannon %>% ggplot(aes(ha_group,value, 
                                                             fill = ha_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(height= 0, width = 0.20, size = 1.5, alpha= 0.4) +
theme_classic(base_size=16)+
  ylab("Shannon Diversity") + xlab("HA Phylogenetic Groups") +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1, size=20), 
        legend.position = "none", axis.text.y = element_text(size=20),
         axis.title.x = element_text(size = 22),  
  axis.title.y = element_text(size = 22)) + 
  scale_fill_manual(values = c(
  "1" = "#56B4E9",
  "2" = "#D55E00",
  "neg" = "#009E73"
))
 
p_formatted_HA_group

```
Statistical analysis of the shannon diversity metric between HA phylogenetic
groups
```{r}
df_HA_group_shannon <- data.frame(ha_group = as.factor(pdata_HA_group_shannon$ha_group), value = pdata_HA_group_shannon$value)
#Kruskal-Wallis and Dunn's test as post-hoc
res_sig_group_HA <-dunn.test(df_HA_group_shannon$value, df_HA_group_shannon$ha_group, kw=TRUE, method = "bh")

view(res_sig_group_HA)

#Two significant groups

```
Generating a boxplot with significance bars. 
```{r}

P_signif_group_HA <- p_formatted_HA_group + geom_signif(
    comparison = list(c("1", "neg"), c("2","neg")),
    y_position = c(4.5,4.25),
    annotations = c("*","*"),
    textsize= 10
  )
P_signif_group_HA

ggsave("Inf_HA_Group_Shannon.png", P_signif_group_HA, width =6, height = 9, dpi = 300)

```
## NA Phylogenetic Groups ##
Shannon diversity calculation but grouped by NA phylogenetic groups. 
```{r}
set.seed(360)
p_NA_group_shannon = plot_richness(psrare, x="na_group",
                        measures = c("Shannon"),
                  color= "na_group")

pdata_NA_group_shannon = p_NA_group_shannon$data

p_formatted_NA_group = pdata_NA_group_shannon %>% ggplot(aes(na_group,value, 
                                                             fill = na_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(height= 0, width = 0.20, size = 1.5, alpha= 0.4) +
theme_classic(base_size=16)+
  ylab("Shannon Diversity") + xlab("NA Phylogenetic Groups") +
theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1, size=20), 
        legend.position = "none", axis.text.y = element_text(size=20),
         axis.title.x = element_text(size = 22),  
  axis.title.y = element_text(size = 22)) + 
 scale_fill_manual(values = c(
  "1" = "darkgrey",
  "2" = "#CC79A7",
  "neg" = "#009E73"
))
p_formatted_NA_group

```
Statistical analysis of the shannon diversity metric between NA phylogenetic
groups
```{r}
df_NA_group_shannon <- data.frame(na_group = as.factor(pdata_NA_group_shannon$na_group), value = pdata_NA_group_shannon$value)
#Kruskal-Wallis and Dunn's test as post-hoc
res_sig_group_NA <-dunn.test(df_NA_group_shannon$value, df_NA_group_shannon$na_group, kw=TRUE, method = "bh")

view(res_sig_group_NA)
```
Generating a boxplot with significance bars. 
```{r}
P_signif_group_NA <- p_formatted_NA_group + geom_signif(
    comparison = list( c("2","neg")),
    y_position = c(4),
    annotations = c("*"),
    textsize = 10
  )
P_signif_group_NA
ggsave("Inf_NA_Group_Shannon.png", P_signif_group_NA, width =6, height = 9, dpi = 300)
```
### Faith's Diversity ###
Calculating the Faith's PD values for each sample.
```{r}

#Adapted from MICB 305
#Taxa converted from columns to rows
otu = as(otu_table(psrare), "matrix")
if(taxa_are_rows(psrare)==TRUE) otu <- t(otu)

# Extract your phyloseq object's tree
tree_faith = phy_tree(psrare)

#Check that all the ASV names match between OTU table and tree
stopifnot(all(colnames(otu) %in% tree_faith$tip.label))

#Remove any unmatched tips
tree_faith = drop.tip(tree_faith, setdiff(tree_faith$tip.label, colnames(otu)))
#Reorder to match
tree_faith = keep.tip(tree_faith, colnames(otu))

#Rooting the tree
if(!is.rooted(tree_faith)) tree_faith <- root(tree_faith, outgroup=tree_faith$tip.label[1], resolve.root=TRUE)

# Calculate Faith's PD for each sample 
#    1. Convert reads to presence/absence and pull into a matrix
otu_pa = otu # Make a copy of your data, then...
otu_pa[otu_pa>0] = 1 #...convert to presence/absence (1 for present, 0 for absent)
#    2. Calculate metric
faith = pd(otu_pa, tree_faith, include.root = TRUE) %>% select(-SR)

#Merging phyloseq data with the Faith's PD data. 

metadata_rarefied = data.frame(sample_data(psrare))

faith_metadata = merge(metadata_rarefied, faith, by= "row.names")

view(faith_metadata)




```
## HA_NA ##
Statistical testing for the Faith's diversity metric between HA_NA subtypes. 
```{r}
faith_HA_NA_df <- data.frame(HA_NA = as.factor(faith_metadata$HA_NA), value = faith_metadata$PD)

faith_HA_NA_sig <-dunn.test(faith_HA_NA_df$value, faith_HA_NA_df$HA_NA, kw=TRUE, method = "bh")
```
## NA Subtype ## 
Statistical testing for the Faith's diversity metric between NA subtypes. 
```{r}
faith_NA_df <- data.frame(na_subtype = as.factor(faith_metadata$na_subtype), 
                                                 value = faith_metadata$PD)
faith_NA_sig <- dunn.test(faith_NA_df$value, faith_NA_df$na_subtype, kw=TRUE, method = "bh")
```
## HA Subtype ## 
Statistical testing for the Faith's diversity metric between HA subtypes. 
```{r}
faith_HA_df <- data.frame(ha_subtype = as.factor(faith_metadata$ha_subtype), 
                                                 value = faith_metadata$PD)
faith_HA_sig <- dunn.test(faith_HA_df$value, faith_HA_df$ha_subtype, kw=TRUE, method = "bh")
```
## NA Subtype Aggregated ##
Statistical testing for the Faith's diversity metric between NA aggregated
subtypes. 
```{r}
faith_NA_aggr_df <- data.frame(na_aggr = as.factor(faith_metadata$na_aggr), 
                                                 value = faith_metadata$PD)
faith_NA_aggr_sig <- dunn.test(faith_NA_aggr_df$value, faith_NA_aggr_df$na_aggr, kw=TRUE, method = "bh")

#Kruskal Wallis of 0.05 but no observed significant pairs
```
## HA Subtype Aggregated ##
Statistical testing for the Faith's diversity metric between HA aggregated
subtypes. 
```{r}
faith_HA_aggr_df <- data.frame(ha_aggr = as.factor(faith_metadata$ha_aggr), 
                                                 value = faith_metadata$PD)
faith_HA_aggr_sig <- dunn.test(faith_HA_aggr_df$value, faith_HA_aggr_df$ha_aggr, kw=TRUE, method = "bh")

#one significant group
```
## HA Phylogenetic Groups ##
Initial boxplot of the Faith's diversity metric grouped by HA phylogenetic 
groups.
```{r}
faith_formatted_HA_group = faith_metadata %>% ggplot(aes(ha_group,PD, 
                                                             fill = ha_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(height= 0, width = 0.20, size = 1.5, alpha= 0.4) +
theme_classic(base_size=16)+
  ylab("Faith's Diversity") + xlab("HA Phylogenetic Groups") +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1, size=20), 
        legend.position = "none", axis.text.y = element_text(size=20),
         axis.title.x = element_text(size = 22),  
  axis.title.y = element_text(size = 22)) + 
   scale_fill_manual(values = c(
  "1" = "#56B4E9",
  "2" = "#D55E00",
  "neg" = "#009E73"
))
faith_formatted_HA_group
```
Statistical testing for the Faith's diversity metric between HA phylogenetic 
groups. 
```{r}
faith_HA_group_df <- data.frame(ha_group = as.factor(faith_metadata$ha_group), 
                                                 value = faith_metadata$PD)
faith_HA_group_sig <- dunn.test(faith_HA_group_df$value, faith_HA_group_df$ha_group, kw=TRUE, method = "bh")

view(faith_HA_group_sig)
```
Significant comparisons mapped onto the initial boxplot.
```{r}
P_faith_group_HA <- faith_formatted_HA_group + geom_signif(
    comparison = list(c("1", "neg"), c("2","neg")),
    y_position = c(16,15.25),
    annotations = c("**","*"),
    textsize = 10
  )
P_faith_group_HA
ggsave("Inf_HA_Group_Faith.png", P_faith_group_HA, width= 6, height =9, dpi=300)
```
## NA Phylogenetic Group ##
Initial boxplot of the Faith's diversity metric grouped by NA phylogenetic 
groups.
```{r}
faith_formatted_NA_group = faith_metadata %>% ggplot(aes(na_group,PD, 
                                                             fill = na_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(height= 0, width = 0.20, size = 1.5, alpha= 0.4) +
theme_classic(base_size=16)+
  ylab("Faith's Diversity") + xlab("NA Phylogenetic Groups") +
   theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1, size=20), 
        legend.position = "none", axis.text.y = element_text(size=20),
         axis.title.x = element_text(size = 22),  
  axis.title.y = element_text(size = 22)) + 
  scale_fill_manual(values = c(
  "1" = "darkgrey",
  "2" = "#CC79A7",
  "neg" = "#009E73"
))
faith_formatted_NA_group
```
Statistical testing for the Faith's diversity metric between NA phylogenetic 
groups. 
```{r}
faith_NA_group_df <- data.frame(na_group = as.factor(faith_metadata$na_group), 
                                                 value = faith_metadata$PD)
faith_NA_group_sig <- dunn.test(faith_NA_group_df$value, faith_NA_group_df$na_group, kw=TRUE, method = "bh")
```
Significant comparisons mapped onto the initial boxplot.
```{r}
P_faith_group_NA <- faith_formatted_NA_group + geom_signif(
    comparison = list(c("1", "neg"), c("2","neg")),
    y_position = c(16,15.25),
    annotations = c("*","**"),
    textsize = 10
  )
P_faith_group_NA
ggsave("Inf_NA_Group_Faith.png", P_faith_group_NA, width= 6, height =9, dpi=300)
```

 
 



