---
title: "alpha_diversity"
output: html_document
author: "Ryan Chen"
---

## Introduction

Alpha diversity of the cloacal microbiomes of influenza A infected mallard ducks was analyzed using Shannon and Faith’s diversity metrics. The code for this analysis was adapted from UBC MICB 305 and MICB 205 course material.

## Load Libraries

```{r}
library(tidyverse)
library(phyloseq)
library(vegan)
library(picante)
library(ape)
library(dunn.test)
library(ggsignif)
```

## Load Data

```{r}
ps <- readRDS('../processed_data/phyloseq_taxonomy.rds')
```

## Rarefy the Data

```{r}
# rarefraction depth was decided based on QIIME2 outputs 
psrare = ps %>% rarefy_even_depth(sample.size = 5396, rngseed = 360)
```

## Shannon's Diversity

### HA Phylogenetic Groups

#### Generate Shannon Metric

```{r}
set.seed(360)
p_HA_group_shannon = plot_richness(psrare, x="ha_group",
                        measures = c("Shannon"),
                  color= "ha_group")

pdata_HA_group_shannon = p_HA_group_shannon$data
```

#### Create initial boxplot (no statistical annotations)

```{r}
p_formatted_HA_group = pdata_HA_group_shannon %>% ggplot(aes(ha_group,value, fill = ha_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(height= 0, width = 0.20, size = 2.5, alpha= 0.4) +
theme_classic(base_size=16)+
  ylab("Shannon Diversity") + xlab("HA Phylogenetic Groups") +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1, size=20), 
        legend.position = "none", axis.text.y = element_text(size=20),
         axis.title.x = element_text(size = 22),  
  axis.title.y = element_text(size = 22)) + 
  scale_fill_manual(values = c(
  "1" = "#56B4E9",
  "2" = "#D55E00",
  "neg" = "#009E73"))
 
p_formatted_HA_group
```

#### Conduct statistical analysis of Shannon diversity between HA groups

```{r}
df_HA_group_shannon <- data.frame(ha_group = as.factor(pdata_HA_group_shannon$ha_group), value = pdata_HA_group_shannon$value)
#Kruskal-Wallis and Dunn's test as post-hoc
res_sig_group_HA <-dunn.test(df_HA_group_shannon$value, df_HA_group_shannon$ha_group, kw=TRUE, method = "bh")

#Two significant groups

```

#### Add statistical annotations to the boxplot and save the figure

```{r}
P_signif_group_HA <- p_formatted_HA_group + geom_signif(
    comparison = list(c("1", "neg"), c("2","neg")),
    y_position = c(4.5,4.25),
    annotations = c("*","*"),
    textsize= 10)
P_signif_group_HA

ggsave("../visualizations/figures/alpha_diversity/HA_Group_Shannon.png",P_signif_group_HA, width = 6, height = 9, dpi = 300)
```

### NA Phylogenetic Groups

#### Generate Shannon Metric

```{r}
set.seed(360)
p_NA_group_shannon = plot_richness(psrare, x="na_group",
                        measures = c("Shannon"),
                  color= "na_group")

pdata_NA_group_shannon = p_NA_group_shannon$data
```

#### Create initial boxplot (no statistical annotations)

```{r}
p_formatted_NA_group = pdata_NA_group_shannon %>% ggplot(aes(na_group,value, fill = na_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(height= 0, width = 0.20, size = 2.5, alpha= 0.4) +
theme_classic(base_size=16)+
  ylab("Shannon Diversity") + xlab("NA Phylogenetic Groups") +
theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1, size=20), 
        legend.position = "none", axis.text.y = element_text(size=20),
         axis.title.x = element_text(size = 22),  
  axis.title.y = element_text(size = 22)) + 
 scale_fill_manual(values = c(
  "1" = "darkgrey",
  "2" = "#CC79A7",
  "neg" = "#009E73"))

p_formatted_NA_group
```

#### Conduct statistical analysis of Shannon diversity between NA groups

```{r}
df_NA_group_shannon <- data.frame(na_group = as.factor(pdata_NA_group_shannon$na_group), value = pdata_NA_group_shannon$value)
#Kruskal-Wallis and Dunn's test as post-hoc
res_sig_group_NA <-dunn.test(df_NA_group_shannon$value, df_NA_group_shannon$na_group, kw=TRUE, method = "bh")
```

#### Add statistical annotations to the boxplot and save the figure

```{r}
P_signif_group_NA <- p_formatted_NA_group + geom_signif(
    comparison = list( c("2","neg")),
    y_position = c(4),
    annotations = c("*"),
    textsize = 10)
P_signif_group_NA

ggsave("../visualizations/figures/alpha_diversity/NA_Group_Shannon.png", P_signif_group_NA, width = 6, height = 9, dpi = 300)
```

## Faith's Diversity

### Calculating Faith's PD values for each sample

```{r}
# Adapted from MICB 305

# Taxa converted from columns to rows
otu = as(otu_table(psrare), "matrix")
if(taxa_are_rows(psrare)==TRUE) otu <- t(otu)

# Extract your phyloseq object's tree
tree_faith = phy_tree(psrare)

# Check that all the ASV names match between OTU table and tree
stopifnot(all(colnames(otu) %in% tree_faith$tip.label))

# Remove any unmatched tips
tree_faith = drop.tip(tree_faith, setdiff(tree_faith$tip.label, colnames(otu)))

# Reorder to match
tree_faith = keep.tip(tree_faith, colnames(otu))

# Rooting the tree
if(!is.rooted(tree_faith)) tree_faith <- root(tree_faith, outgroup=tree_faith$tip.label[1], resolve.root=TRUE)

# Calculate Faith's PD for each sample 
#    1. Convert reads to presence/absence and pull into a matrix

otu_pa = otu # Make a copy of your data, then...
otu_pa[otu_pa>0] = 1 #...convert to presence/absence (1 for present, 0 for absent)

#    2. Calculate metric
faith = pd(otu_pa, tree_faith, include.root = TRUE) %>% select(-SR)

# Merging phyloseq data with the Faith's PD data. 
metadata_rarefied = data.frame(sample_data(psrare))

faith_metadata = merge(metadata_rarefied, faith, by= "row.names")
```

### HA Phylogenetic Groups

#### Create initial boxplot (no statistical annotations)

```{r}
faith_formatted_HA_group = faith_metadata %>% ggplot(aes(ha_group,PD, fill = ha_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(height= 0, width = 0.20, size = 2.5, alpha= 0.4) +
theme_classic(base_size=16)+
  ylab("Faith's Diversity") + xlab("HA Phylogenetic Groups") +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1, size=20), 
        legend.position = "none", axis.text.y = element_text(size=20),
         axis.title.x = element_text(size = 22),  
  axis.title.y = element_text(size = 22)) + 
   scale_fill_manual(values = c(
  "1" = "#56B4E9",
  "2" = "#D55E00",
  "neg" = "#009E73"))

faith_formatted_HA_group
```

#### Conduct statistical analysis of Faith's diversity between HA groups

```{r}
faith_HA_group_df <- data.frame(ha_group = as.factor(faith_metadata$ha_group), value = faith_metadata$PD)
faith_HA_group_sig <- dunn.test(faith_HA_group_df$value, faith_HA_group_df$ha_group, kw=TRUE, method = "bh")
```

#### Add statistical annotations to the boxplot and save the figure

```{r}
P_faith_group_HA <- faith_formatted_HA_group + geom_signif(
    comparison = list(c("1", "neg"), c("2","neg")),
    y_position = c(16,15.25),
    annotations = c("**","*"),
    textsize = 10)

P_faith_group_HA

ggsave("../visualizations/figures/alpha_diversity/HA_Group_Faith.png", P_faith_group_HA, width = 6, height =9, dpi=300)
```

### NA Phylogenetic Group

#### Create initial boxplot (no statistical annotations)

```{r}
faith_formatted_NA_group = faith_metadata %>% ggplot(aes(na_group,PD, fill = na_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(height= 0, width = 0.20, size = 2.5, alpha= 0.4) +
theme_classic(base_size=16)+
  ylab("Faith's Diversity") + xlab("NA Phylogenetic Groups") +
   theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1, size=20), 
        legend.position = "none", axis.text.y = element_text(size=20),
         axis.title.x = element_text(size = 22),  
  axis.title.y = element_text(size = 22)) + 
  scale_fill_manual(values = c(
  "1" = "darkgrey",
  "2" = "#CC79A7",
  "neg" = "#009E73"))

faith_formatted_NA_group
```

#### Conduct statistical analysis of Faith's diversity between NA groups

```{r}
faith_NA_group_df <- data.frame(na_group = as.factor(faith_metadata$na_group), value = faith_metadata$PD)
faith_NA_group_sig <- dunn.test(faith_NA_group_df$value, faith_NA_group_df$na_group, kw=TRUE, method = "bh")
```

#### Add statistical annotations to the boxplot and save the figure

```{r}
P_faith_group_NA <- faith_formatted_NA_group + geom_signif(
    comparison = list(c("1", "neg"), c("2","neg")),
    y_position = c(16,15.25),
    annotations = c("*","**"),
    textsize = 10)

P_faith_group_NA

ggsave("../visualizations/figures/alpha_diversity/NA_Group_Faith.png", P_faith_group_NA, width = 6, height =9, dpi=300)
```
