---
title: "Beta Diversity Final project"
author: "Soroush_Torfehnejad"
date: "2026-03-29"
output: html_document
---

# Step 1)

### Load libraries and Data, I rarefied the data based on the conclusions we came to when looking at the QUIME results.The rarefaction number is 5396

```{r Load data and libraries}
library(tidyverse)
library(phyloseq)
library(vegan)
# Load object
ps <- readRDS('../processed_data/phyloseq_taxonomy.rds')


# rarefy the data:
psrare <- ps|>
          rarefy_even_depth(sample.size = 5396, rngseed = 421)
table(sample_sums(psrare))


```

#Step 2) 
### Caclculate the beta diversity using the Bray-Curtis distance matrix.

```{r calculate bray-curtis distances}
# Calculate the distance between each sample pair
ps_WUF <- phyloseq::distance(psrare, method = "wunifrac")
View(as.matrix(ps_WUF)) # Every pairwise comparison included
```

#Step 3) 
### use Multidimensional scaling to reduce the distance matrix into fewer dimentions which will help plot abd understand the Data.

```{r normalize and plot}
# MDS scaling
set.seed(421)
mds = metaMDS(ps_WUF)

# Extract data
mds_data = mds$points %>% as.data.frame %>%
  merge(sample_data(psrare), by='row.names', sort=F)
head(mds_data)
```

#Step 4) 
### Plot the data based on the following comparisons, different HA/NA subtypes, different HA subtypes, different NA subtypes, finally Infected Vs non infection comparison.

```{r}

# looking at Beta diversity differences between HA group 1, group 2 and negative control

plot_HA_G1_VS_G2_VS_neg<- mds_data %>%
                  ggplot(aes(MDS1,MDS2,color = ha_group)) +
                  geom_point() +
                  stat_ellipse() +
                  theme_classic(base_size=18)



# looking at Beta diversity differences between HA group 1 and group 2
plot_HA_G1_VS_G2<- mds_data %>%
                  filter(ha_group != "neg") %>% 
                  ggplot(aes(MDS1,MDS2,color = ha_group)) +
                  geom_point() +
                  stat_ellipse() +
                  theme_classic(base_size=18)


# looking at Beta diversity differences between HA group 1 and uninfected
plot_HA_G1_VS_NEG<- mds_data %>%
                  filter(ha_group != "2") %>% 
                  ggplot(aes(MDS1,MDS2,color = ha_group)) +
                  geom_point() +
                  stat_ellipse() +
                  theme_classic(base_size=18)

# looking at Beta diversity differences between HA group 2 and uninfected
plot_HA_G2_VS_NEG<- mds_data %>%
                  filter(ha_group != "1") %>% 
                  ggplot(aes(MDS1,MDS2,color = ha_group)) +
                  geom_point() +
                  stat_ellipse() +
                  theme_classic(base_size=18)

plot_HA_G1_VS_G2_VS_neg

plot_HA_G1_VS_G2

plot_HA_G1_VS_NEG

plot_HA_G2_VS_NEG
```

#Step 5)

### Perform a PERMANOVA test to find statistical significance between the groups using an Adonis function.

```{r}
metadata <- sample_data(psrare) %>% data.frame() # to make the next lines easier


set.seed(421)
# Single variable anova 

Permanova_on_HA_groups <- adonis2(ps_WUF ~ ha_group, data = metadata, permutations = 99999)

Permanova_on_HA_groups

# when observing if their is a difference in the Beta diversity between Group 1,2 or negative control statistical significance was seen p= 0.02326


#this statistical significance led to us comparing each individual group within the ha_group column: group1 vs group 2 vs uninfected


# Comparing HA G1 Vs G2 using PERMANOVA 
compare_G1_vs_G2 = c('2','1')

# Correct the data to make them have the same size when comparing G1 Vs G2
keep_G1_vs_G2 = metadata %>% filter(ha_group %in% compare_G1_vs_G2) %>% rownames()


ps_WUF_G1_vs_G2 = as.matrix(ps_WUF)[keep_G1_vs_G2,keep_G1_vs_G2] %>% as.dist()

# Run a PERMANOVA on G1 vs G2
Permanova_G1_vs_G2 = adonis2(ps_WUF_G1_vs_G2 ~ ha_group,
                       data = metadata %>% filter(ha_group %in% compare_G1_vs_G2), permutations = 99999)

Permanova_G1_vs_G2 

# This one showed no significance p= 0.3286 indcation no differences between the beta diversity of G1 vs G2

# Comparing HA G1 Vs Neg using PERMANOVA 
compare_G1_vs_neg = c('neg','1')

# Correct the data to make them have the same size when comparing G1 Vs neg
keep_G1_vs_neg = metadata %>% filter(ha_group %in% compare_G1_vs_neg) %>% rownames()


ps_WUF_G1_vs_neg = as.matrix(ps_WUF)[keep_G1_vs_neg,keep_G1_vs_neg] %>% as.dist()

# Run a PERMANOVA on G1 vs neg
Permanova_G1_vs_neg = adonis2(ps_WUF_G1_vs_neg ~ ha_group,
                       data = metadata %>% filter(ha_group %in% compare_G1_vs_neg), permutations = 99999)

Permanova_G1_vs_neg 


# This one showed no significance p= 0.09864 indcation no differences between the beta diversity of G1 vs neg


# Comparing HA G2 Vs Neg using PERMANOVA 
compare_G2_vs_neg = c('neg','2')

# Correct the data to make them have the same size when comparing G2 Vs neg
keep_G2_vs_neg = metadata %>% filter(ha_group %in% compare_G2_vs_neg) %>% rownames()


ps_WUF_G2_vs_neg = as.matrix(ps_WUF)[keep_G2_vs_neg,keep_G2_vs_neg] %>% as.dist()

# Run a PERMANOVA on G2 vs neg
Permanova_G2_vs_neg = adonis2(ps_WUF_G2_vs_neg ~ ha_group,
                       data = metadata %>% filter(ha_group %in% compare_G2_vs_neg), permutations = 99999)

Permanova_G2_vs_neg 

# This one showed no significance p= 0.0142 indcation a statisticaly significant differences between the beta diversity of G2 vs neg. This indicates that group 2 HA influenza viruses may have a stronger affect in changing the beta diversity of the coacle microbiome. 


```
