---
title: "beta_diversity"
output: html_document
author: "Soroush Torfehnejad"
---

# Beta-Diversity Analysis

## Step 1. Load Libraries and Data

The rarefaction depth of 5396 was selected based on results obtained during the QIIME2 processing steps.

```{r Load data, libraries, and subcatagoriuze phylosq tables}
library(tidyverse)
library(phyloseq)
library(vegan)

# Load data
ps <- readRDS('../processed_data/phyloseq_taxonomy.rds')

# Rarefy data
psrare <- ps|>
          rarefy_even_depth(sample.size = 5396, rngseed = 421)
table(sample_sums(psrare))

```

### Step 1.5. Create 8 Phyloseq Objects Filtered Appropriately for each Comparison

```{r}
# HA comparisons
ps_ha_g1_g2 = psrare
ps_ha_g1_g2@sam_data <- data.frame(ps_ha_g1_g2@sam_data) %>% 
  filter(ps_ha_g1_g2@sam_data$ha_group != "neg") %>%
  sample_data(.)

ps_ha_g1_neg = psrare
ps_ha_g1_neg@sam_data <- data.frame(ps_ha_g1_neg@sam_data) %>% 
  filter(ps_ha_g1_neg@sam_data$ha_group != "2") %>%
  sample_data(.)

ps_ha_g2_neg = psrare
ps_ha_g2_neg@sam_data <- data.frame(ps_ha_g2_neg@sam_data) %>% 
  filter(ps_ha_g2_neg@sam_data$ha_group != "1") %>%
  sample_data(.)


# NA comparisons 
ps_na_g1_g2 = psrare
ps_na_g1_g2@sam_data <- data.frame(ps_na_g1_g2@sam_data) %>% 
  filter(ps_na_g1_g2@sam_data$na_group != "neg") %>%
  sample_data(.)

ps_na_g1_neg = psrare
ps_na_g1_neg@sam_data <- data.frame(ps_na_g1_neg@sam_data) %>% 
  filter(ps_na_g1_neg@sam_data$na_group != "2") %>%
  sample_data(.)

ps_na_g2_neg = psrare
ps_na_g2_neg@sam_data <- data.frame(ps_na_g2_neg@sam_data) %>% 
  filter(ps_na_g2_neg@sam_data$na_group != "1") %>%
  sample_data(.)
```

## Step 2. Caclculate Beta-Diversity Using Weighted Unifrac Distance

```{r calculate wighted unifrac distances}
# Calculate the distance between each sample pair HA and NA phylogeny groups
ps_all_groups_WUF <- phyloseq::distance(psrare, method = "wunifrac")

# HA Group 1 vs. Group 2 weighted unifrac distances. 
ps_ha_g1_g2_WUF <- phyloseq::distance(ps_ha_g1_g2, method = "wunifrac")

# HA Group 1 vs. neg weighted unifrac distances. 
ps_ha_g1_neg_WUF <- phyloseq::distance(ps_ha_g1_neg, method = "wunifrac")

# HA Group 2 vs. neg weighted unifrac distances. 
ps_ha_g2_neg_WUF <- phyloseq::distance(ps_ha_g2_neg, method = "wunifrac")

# NA Group 1 vs. Group 2 weighted unifrac distances. 
ps_na_g1_g2_WUF <- phyloseq::distance(ps_na_g1_g2, method = "wunifrac")

# NA Group 1 vs. neg weighted unifrac distances. 
ps_na_g1_neg_WUF <- phyloseq::distance(ps_na_g1_neg, method = "wunifrac")

# NA Group 2 vs. neg weighted unifrac distances. 
ps_na_g2_neg_WUF <- phyloseq::distance(ps_na_g2_neg, method = "wunifrac")
```

## Step 3. Use Multidimensional Scaling (MDS) to Reduce Distance Matrix Dimensionality (and assist in data interpretation/plotting).

### Conduct MDS

```{r normalize and plot}
# set a seed
set.seed(421)

# MDS scaling for HA and NA all groups 
mds_all_groups = metaMDS(ps_all_groups_WUF)

# MDS scaling for HA G1 Vs G2
mds_ha_g1_g2 = metaMDS(ps_ha_g1_g2_WUF)

# MDS scaling for HA G1 Vs neg
mds_ha_g1_neg = metaMDS(ps_ha_g1_neg_WUF)

# MDS scaling for HA G2 Vs neg
mds_ha_g2_neg = metaMDS(ps_ha_g2_neg_WUF)

# MDS scaling for NA G1 Vs G2
mds_na_g1_g2 = metaMDS(ps_na_g1_g2_WUF)

# MDS scaling for NA G1 Vs neg
mds_na_g1_neg = metaMDS(ps_na_g1_neg_WUF)

# MDS scaling for NA G2 Vs neg
mds_na_g2_neg = metaMDS(ps_na_g2_neg_WUF)
```

### Extract MDS Data

```{r}
# MDS scaling for HA and NA all groups 
mds_all_groups_data = mds_all_groups$points %>% as.data.frame %>%
  merge(sample_data(psrare), by='row.names', sort=F)

# MDS scaling for HA G1 Vs G2
mds_ha_g1_g2_data = mds_ha_g1_g2$points %>% as.data.frame %>%
  merge(sample_data(ps_ha_g1_g2), by='row.names', sort=F)

# MDS scaling for HA G1 Vs neg
mds_ha_g1_neg_data = mds_ha_g1_neg$points %>% as.data.frame %>%
  merge(sample_data(ps_ha_g1_neg), by='row.names', sort=F)

# MDS scaling for HA G2 Vs neg
mds_ha_g2_neg_data = mds_ha_g2_neg$points %>% as.data.frame %>%
  merge(sample_data(ps_ha_g2_neg), by='row.names', sort=F)

# MDS scaling for NA G1 Vs G2
mds_na_g1_g2_data = mds_na_g1_g2$points %>% as.data.frame %>%
  merge(sample_data(ps_na_g1_g2), by='row.names', sort=F)

# MDS scaling for NA G1 Vs neg
mds_na_g1_neg_data = mds_na_g1_neg$points %>% as.data.frame %>%
  merge(sample_data(ps_na_g1_neg), by='row.names', sort=F)

# MDS scaling for NA G2 Vs neg
mds_na_g2_neg_data = mds_na_g2_neg$points %>% as.data.frame %>%
  merge(sample_data(ps_na_g2_neg), by='row.names', sort=F)
```

## Step 4. Plot the HA Data

Data will be plot according to the following comparisons: - HA G1 vs. G2 vs. neg (uninfected samples) - HA G1 vs. G2 - HA G1 vs. neg (uninfected samples) - HA G2 vs. neg (uninfected samples)

### Standardization of plot shapes/colours

```{r}
# Standardizing all the colors and shapes so they are consistent when comparing graphs. 
custom_colors_HA<-c("1" = "#56B4E9", "2" = "#D55E00", "neg" = "#009E73")
custom_shapes_HA <- c("1" = 21, "2" = 24, "neg" = 22)
```

### HA G1 vs. G2 vs. neg

```{r}
plot_HA_G1_VS_G2_VS_neg<- mds_all_groups_data %>%
                                  ggplot(aes(MDS1,MDS2,color = ha_group, fill = ha_group, shape = ha_group )) +
                                  geom_point(size = 1.5, stroke = 1.4) +
                                  scale_color_manual(values = custom_colors_HA)+
                                  scale_fill_manual(values = custom_colors_HA)+
                                  scale_shape_manual(values = custom_shapes_HA)+
                                  stat_ellipse() +
                                  theme_classic(base_size=18)+
                                  theme(axis.title = element_text(size = 12),
                                        axis.text = element_text(size = 10))+
                                  labs( x = "MDS Axis 1",
                                        y = "MDS Axis 2",
                                        title = "Beta Diversity All HA Groups",
                                        color = "HA grouping",
                                        fill = "HA grouping",
                                        shape = "HA grouping")
                           
plot_HA_G1_VS_G2_VS_neg            
```

### HA G1 vs. G2

```{r}
plot_HA_G1_VS_G2<- mds_ha_g1_g2_data %>%
                            ggplot(aes(MDS1,MDS2,color = ha_group, fill = ha_group, shape = ha_group )) +
                            geom_point(size = 1.5, stroke = 1.4) +
                            scale_color_manual(values = custom_colors_HA)+
                            scale_fill_manual(values = custom_colors_HA)+
                            scale_shape_manual(values = custom_shapes_HA)+
                            stat_ellipse() +
                            theme_classic(base_size=18)+
                            theme(axis.title = element_text(size = 12),
                                  axis.text = element_text(size = 10))+
                            labs( x = "MDS Axis 1",
                                  y = "MDS Axis 2",
                                  title = "Beta Diversity HA Group 1 vs Group 2",
                                  color = "HA grouping",
                                  fill = "HA grouping",
                                  shape = "HA grouping")
                  
plot_HA_G1_VS_G2
```

### HA G1 vs. neg

```{r}
plot_HA_G1_VS_NEG<- mds_ha_g1_neg_data %>%
                              ggplot(aes(MDS1,MDS2,color = ha_group, fill = ha_group, shape = ha_group )) +
                              geom_point(size = 1.5, stroke = 1.4) +
                              scale_color_manual(values = custom_colors_HA)+
                              scale_fill_manual(values = custom_colors_HA)+
                              scale_shape_manual(values = custom_shapes_HA)+
                              stat_ellipse() +
                              theme_classic(base_size=18)+
                              theme(axis.title = element_text(size = 12),
                                    axis.text = element_text(size = 10))+
                              labs( x = "MDS Axis 1",
                                    y = "MDS Axis 2",
                                    title = "Beta Diversity HA Group 1 Vs Uninfected",
                                    color = "HA grouping",
                                    fill = "HA grouping",
                                    shape = "HA grouping")

plot_HA_G1_VS_NEG
```

### HA G2 vs. neg

```{r}
plot_HA_G2_VS_NEG<- mds_ha_g2_neg_data %>%
                              ggplot(aes(MDS1,MDS2,color = ha_group, fill = ha_group, shape = ha_group )) +
                              geom_point(size = 1.5, stroke = 1.4) +
                              scale_color_manual(values = custom_colors_HA)+
                              scale_fill_manual(values = custom_colors_HA)+
                              scale_shape_manual(values = custom_shapes_HA)+
                              stat_ellipse() +
                              theme_classic(base_size=18)+
                              theme(axis.title = element_text(size = 12),
                                    axis.text = element_text(size = 10))+
                              labs( x = "MDS Axis 1",
                                    y = "MDS Axis 2",
                                    title = "Beta Diversity HA Group 2 Vs Uninfected",
                                    color = "HA grouping",
                                    fill = "HA grouping",
                                    shape = "HA grouping")
plot_HA_G2_VS_NEG
```

## Step 5. Perform PERMANOVA testing on HA Comparisons

PERMANOVA tests are used to look for statistical significance between HA groups (and uninfected samples) using the adonis2 function.

```{r}
# Make metadata for all comparisons 
metadata_all_groups <- sample_data(psrare) %>% data.frame() # all groups 
metadata_ha_g1_g2 <- sample_data(ps_ha_g1_g2) %>% data.frame() # HA G1 vs G2
metadata_ha_g1_neg <- sample_data(ps_ha_g1_neg) %>% data.frame() # HA G1 vs Neg
metadata_ha_g2_neg <- sample_data(ps_ha_g2_neg) %>% data.frame() # HA G2 vs Neg

# Set seed
set.seed(421)


# Single variable anova 
Permanova_on_HA_groups <- adonis2(ps_all_groups_WUF ~ ha_group, data = metadata_all_groups, permutations = 99999)
Permanova_on_HA_groups

# When observing if their is a difference in the beta-diversity between HA Group 1, 2 or negative control, statistical significance was seen (p = 0.02326). This prompted comparing each individual group within the ha_group column: group1 vs group 2 vs uninfected


# Comparing HA G1 vs. G2 using PERMANOVA 
compare_G1_vs_G2_HA <- c('2','1')

# Correct the data to make them have the same size when comparing G1 vs. G2
keep_G1_vs_G2_HA <- metadata_ha_g1_g2 %>% filter(ha_group %in% compare_G1_vs_G2_HA) %>% rownames()

ps_WUF_G1_vs_G2_HA <- as.matrix(ps_ha_g1_g2_WUF)[keep_G1_vs_G2_HA,keep_G1_vs_G2_HA] %>% as.dist()

# Run a PERMANOVA on G1 vs. G2
Permanova_HA_G1_vs_G2 <- adonis2(ps_WUF_G1_vs_G2_HA ~ ha_group,
                       data <- metadata_ha_g1_g2 %>% filter(ha_group %in% compare_G1_vs_G2_HA),permutations = 99999)
Permanova_HA_G1_vs_G2 

# This comparison showed no significance (p = 0.3286), indicative of no differences between the beta-diversity of G1 vs G2


# Comparing HA G1 vs Neg using PERMANOVA 
compare_G1_vs_neg_HA <- c('neg','1')

# Correct the data to make them have the same size when comparing G1 Vs neg
keep_G1_vs_neg_HA <- metadata_ha_g1_neg %>% filter(ha_group %in% compare_G1_vs_neg_HA) %>% rownames()
ps_WUF_G1_vs_neg_HA <- as.matrix(ps_ha_g1_neg_WUF)[keep_G1_vs_neg_HA,keep_G1_vs_neg_HA] %>% as.dist()

# Run a PERMANOVA on G1 vs neg
Permanova_HA_G1_vs_neg <- adonis2(ps_WUF_G1_vs_neg_HA ~ ha_group,
                       data = metadata_ha_g1_neg %>% filter(ha_group %in% compare_G1_vs_neg_HA), permutations = 99999)
Permanova_HA_G1_vs_neg 

# This comparison showed no significance (p = 0.09864), indicative of no differences between the beta-diversity of G1 vs neg


# Comparing HA G2 Vs Neg using PERMANOVA 
compare_G2_vs_neg_HA <- c('neg','2')

# Correct the data to make them have the same size when comparing G2 Vs neg
keep_G2_vs_neg_HA <- metadata_ha_g2_neg %>% filter(ha_group %in% compare_G2_vs_neg_HA) %>% rownames()
ps_WUF_G2_vs_neg_HA <- as.matrix(ps_ha_g2_neg_WUF)[keep_G2_vs_neg_HA,keep_G2_vs_neg_HA] %>% as.dist()

# Run a PERMANOVA on G2 vs neg
Permanova_HA_G2_vs_neg <- adonis2(ps_WUF_G2_vs_neg_HA ~ ha_group,
                       data = metadata_ha_g2_neg %>% filter(ha_group %in% compare_G2_vs_neg_HA), permutations = 99999)
Permanova_HA_G2_vs_neg 

# This comparison showed significance (p = 0.01333), indicative of statistically significant differences between the beta-diversity of G2 vs neg. That may indicate that HA group 2 has a stronger effect on the beta-diversity of the cloacal microbiome.
```

### Step 5.5. Adjust p-values to account for multiple comparisons

```{r}
# Adjusted P values using the Benjamini-Hochberg procedure
p_HA<-c(0.02326,0.3286,0.09864,0.01333)
p.adjust(p_HA, method = "BH")

# the adjusted P values are as follows:
  # G1 vs G2 vs NEG = 0.04652
  # G1 vs G2 = 0.32860 
  # G1 vs neg = 0.13152 
  # G2 vs neg = 0.04652
```

## Step 6. Plot the NA Data

Data will be plot according to the following comparisons: - NA G1 vs. G2 vs. neg (uninfected samples) - NA G1 vs. G2 - NA G1 vs. neg (uninfected samples) - NA G2 vs. neg (uninfected samples)

### Standardization of plot shapes/colours

```{r}
# Standardizing all the colors and shapes so they are consistent when comparing graphs. 
custom_colors_NA<-c("1" = "darkgrey", "2" = "#CC79A7", "neg" = "#009E73")
custom_shapes_NA <- c("1" = 21, "2" = 24, "neg" = 22)
```

### NA G1 vs. G2 vs. neg

```{r}
plot_NA_G1_VS_G2_VS_neg<- mds_all_groups_data %>%
                                  ggplot(aes(MDS1,MDS2,color = na_group, fill = na_group, shape = na_group )) +
                                  geom_point(size = 1.5, stroke = 1.4) +
                                  scale_color_manual(values = custom_colors_NA)+
                                  scale_fill_manual(values = custom_colors_NA)+
                                  scale_shape_manual(values = custom_shapes_NA)+
                                  stat_ellipse() +
                                  theme_classic(base_size=18)+
                                  theme(axis.title = element_text(size = 12),
                                        axis.text = element_text(size = 10))+
                                  labs( x = "MDS Axis 1",
                                        y = "MDS Axis 2",
                                        title = "Beta Diversity All NA Groups",
                                        color = "NA grouping",
                                        fill = "NA grouping",
                                        shape = "NA grouping")
        
plot_NA_G1_VS_G2_VS_neg
```

### NA G1 vs. G2

```{r}
plot_NA_G1_VS_G2<- mds_na_g1_g2_data %>%
                            ggplot(aes(MDS1,MDS2,color = na_group, fill = na_group, shape = na_group )) +
                            geom_point(size = 1.5, stroke = 1.4) +
                            scale_color_manual(values = custom_colors_NA)+
                            scale_fill_manual(values = custom_colors_NA)+
                            scale_shape_manual(values = custom_shapes_NA)+
                            stat_ellipse() +
                            theme_classic(base_size=18)+
                            theme(axis.title = element_text(size = 12),
                                  axis.text = element_text(size = 10))+
                            labs( x = "MDS Axis 1",
                                  y = "MDS Axis 2",
                                  title = "Beta Diversity NA Group 1 vs Group 2",
                                  color = "NA grouping",
                                  fill = "NA grouping",
                                  shape = "NA grouping")
                  
plot_NA_G1_VS_G2
```

### NA G1 vs. neg

```{r}
plot_NA_G1_VS_NEG<- mds_na_g1_neg_data %>%
                              ggplot(aes(MDS1,MDS2,color = na_group, fill = na_group, shape = na_group )) +
                              geom_point(size = 1.5, stroke = 1.4) +
                              scale_color_manual(values = custom_colors_NA)+
                              scale_fill_manual(values = custom_colors_NA)+
                              scale_shape_manual(values = custom_shapes_NA)+
                              stat_ellipse() +
                              theme_classic(base_size=18)+
                              theme(axis.title = element_text(size = 12),
                                    axis.text = element_text(size = 10))+
                              labs( x = "MDS Axis 1",
                                    y = "MDS Axis 2",
                                    title = "Beta Diversity NA G1 Vs Uninfected",
                                    color = "NA grouping",
                                    fill = "NA grouping",
                                    shape = "NA grouping")

plot_NA_G1_VS_NEG
```

### NA G2 vs. neg

```{r}
plot_NA_G2_VS_NEG<- mds_na_g2_neg_data %>%
                              ggplot(aes(MDS1,MDS2,color = na_group, fill = na_group, shape = na_group )) +
                              geom_point(size = 1.5, stroke = 1.4) +
                              scale_color_manual(values = custom_colors_NA)+
                              scale_fill_manual(values = custom_colors_NA)+
                              scale_shape_manual(values = custom_shapes_NA)+
                              stat_ellipse() +
                              theme_classic(base_size=18)+
                              theme(axis.title = element_text(size = 12),
                                    axis.text = element_text(size = 10))+
                              labs( x = "MDS Axis 1",
                                    y = "MDS Axis 2",
                                    title = "Beta Diversity NA Group 2 Vs Uninfected",
                                    color = "NA grouping",
                                    fill = "NA grouping",
                                    shape = "NA grouping")
plot_NA_G2_VS_NEG
```

## Step 7. Perform PERMANOVA testing on NA Comparisons

```{r}
# Make metadata for all comparisons 
metadata_all_groups <- sample_data(psrare) %>% data.frame() # all groups 
metadata_na_g1_g2 <- sample_data(ps_na_g1_g2) %>% data.frame() # HA G1 vs G2
metadata_na_g1_neg <- sample_data(ps_na_g1_neg) %>% data.frame() # HA G1 vs Neg
metadata_na_g2_neg <- sample_data(ps_na_g2_neg) %>% data.frame() # HA G2 vs Neg

# Set seed
set.seed(421)


# Single variable anova
Permanova_on_NA_groups <- adonis2(ps_all_groups_WUF ~ na_group, data = metadata_all_groups, permutation = 99999)
Permanova_on_NA_groups
# @hen observing if their is a difference in the beta-diversity between NA Group 1, 2 or negative control, statistical significance was seen (p = 0.0235). This statistical significance led to us comparing each individual group within the na_group column: group 1 vs group 2 vs uninfected.


# Comparing NA G1 Vs G2 using PERMANOVA 
compare_G1_vs_G2_NA <- c('2','1')

# Correct the data to make them have the same size when comparing G1 Vs G2
keep_G1_vs_G2_NA <- metadata_na_g1_g2 %>% filter(na_group %in% compare_G1_vs_G2_NA) %>% rownames()
ps_WUF_G1_vs_G2_NA <- as.matrix(ps_na_g1_g2_WUF)[keep_G1_vs_G2_NA,keep_G1_vs_G2_NA] %>% as.dist()

# Run a PERMANOVA on G1 vs G2
Permanova_NA_G1_vs_G2 <- adonis2(ps_WUF_G1_vs_G2_NA ~ na_group,
                       data = metadata_na_g1_g2 %>% filter(na_group %in% compare_G1_vs_G2_NA), permutation = 99999)
Permanova_NA_G1_vs_G2 

# This comparison showed no significance (p = 0.2812), indicative of no differences between the beta-diversity of NA G1 vs G2.


# Comparing HA G1 Vs Neg using PERMANOVA 
compare_G1_vs_neg_NA <- c('neg','1')

# Correct the data to make them have the same size when comparing G1 Vs neg
keep_G1_vs_neg_NA <- metadata_na_g1_neg %>% filter(ha_group %in% compare_G1_vs_neg_NA) %>% rownames()
ps_WUF_G1_vs_neg_NA <- as.matrix(ps_na_g1_neg_WUF)[keep_G1_vs_neg_NA,keep_G1_vs_neg_NA] %>% as.dist()

# Run a PERMANOVA on G1 vs neg
Permanova_NA_G1_vs_neg <- adonis2(ps_WUF_G1_vs_neg_NA ~ na_group,
                       data = metadata_na_g1_neg %>% filter(ha_group %in% compare_G1_vs_neg_NA), permutations = 99999)
Permanova_NA_G1_vs_neg 

# This comparison showed no significance (p = 0.3379), indicative of no differences between the beta-diversity of NA G1 vs neg.


# Comparing HA G2 Vs Neg using PERMANOVA 
compare_G2_vs_neg_NA <- c('neg','2')

# Correct the data to make them have the same size when comparing G2 Vs neg
keep_G2_vs_neg_NA <- metadata_na_g2_neg %>% filter(ha_group %in% compare_G2_vs_neg_NA) %>% rownames()
ps_WUF_G2_vs_neg_NA <- as.matrix(ps_na_g2_neg_WUF)[keep_G2_vs_neg_NA,keep_G2_vs_neg_NA] %>% as.dist()

# Run a PERMANOVA on G2 vs neg
Permanova_NA_G2_vs_neg <- adonis2(ps_WUF_G2_vs_neg_NA ~ na_group,
                       data = metadata_na_g2_neg %>% filter(ha_group %in% compare_G2_vs_neg_NA), permutations = 99999)
Permanova_NA_G2_vs_neg 

# This comparison showed significance (p = 0.04059), an indcation of statistically significant differences between the beta-diversity of G2 vs neg. This may indicate that NA group 2 has a stronger effect on the beta-diversity of the cloacal microbiome.
```

### Step 7.5. Adjust p-values to account for multiple comparisons

```{r}
# Adjusted P values using the Benjamini-Hochberg procedure
p_NA<-c(0.0235,0.2812,0.3379,0.04059)
p.adjust(p_NA, method = "BH")

# the adjusted P values are as follows:
  # G1 vs G2 vs neg = 0.08118
  # G1 vs G2 = 0.33790 
  # G1 vs neg = 0.33790 
  # G2 vs neg = 0.08118
```

## Step 8. Save the Plots and Data

### Plots

```{r}
# HA plots
ggsave('../visualizations/figures/beta_diversity/Beta Diversity HA G1 vs G2 vs NEG.jpeg',
       plot = plot_HA_G1_VS_G2_VS_neg,
       height= 4, width = 6.5)
ggsave('../visualizations/figures/beta_diversity/Beta Diversity HA G1 vs G2.jpeg',
       plot = plot_HA_G1_VS_G2,
       height= 4, width = 6.5)
ggsave('../visualizations/figures/beta_diversity/Beta Diversity HA G1 vs NEG.jpeg',
       plot = plot_HA_G1_VS_NEG,
       height= 4, width = 6.5)
ggsave('../visualizations/figures/beta_diversity/Beta Diversity HA G2 vs NEG.jpeg',
       plot = plot_HA_G2_VS_NEG,
       height= 4, width = 6.5)

# NA plots
ggsave('../visualizations/figures/beta_diversity/Beta Diversity NA G1 vs G2 vs NEG.jpeg',
       plot = plot_NA_G1_VS_G2_VS_neg,
       height= 4, width = 6.5)
ggsave('../visualizations/figures/beta_diversity/Beta Diversity NA G1 vs G2.jpeg',
       plot = plot_NA_G1_VS_G2,
       height= 4, width = 6.5)
ggsave('../visualizations/figures/beta_diversity/Beta Diversity NA G1 vs NEG.jpeg',
       plot = plot_NA_G1_VS_NEG,
       height= 4, width = 6.5)
ggsave('../visualizations/figures/beta_diversity/Beta Diversity NA G2 vs NEG.jpeg',
       plot = plot_NA_G2_VS_NEG,
       height= 4, width = 6.5)
```

### Data (save statistics within Excel files)

```{r}
# HA Comparisons
writexl::write_xlsx(list('Beta Diversity' = Permanova_on_HA_groups),
                    '../visualizations/tables/beta_diversity/Beta Diversity Permanova_on_HA_groups.xlsx')
writexl::write_xlsx(list('Beta Diversity' = Permanova_HA_G1_vs_G2),
                    '../visualizations/tables/beta_diversity/Beta Diversity Permanova_HA_G1_vs_G2.xlsx')
writexl::write_xlsx(list('Beta Diversity' = Permanova_HA_G1_vs_neg),
                    '../visualizations/tables/beta_diversity/Beta Diversity Permanova_HA_G1_vs_neg.xlsx')
writexl::write_xlsx(list('Beta Diversity' = Permanova_HA_G2_vs_neg),
                    '../visualizations/tables/beta_diversity/Beta Diversity Permanova_HA_G2_vs_neg.xlsx')

# NA Comparisons
writexl::write_xlsx(list('Beta Diversity' = Permanova_on_NA_groups),
                    '../visualizations/tables/beta_diversity/Beta Diversity Permanova_on_NA_groups.xlsx')
writexl::write_xlsx(list('Beta Diversity' = Permanova_NA_G1_vs_G2),
                    '../visualizations/tables/beta_diversity/Beta Diversity Permanova_NA_G1_vs_G2.xlsx')
writexl::write_xlsx(list('Beta Diversity' = Permanova_NA_G1_vs_neg),
                    '../visualizations/tables/beta_diversity/Beta Diversity Permanova_NA_G1_vs_neg.xlsx')
writexl::write_xlsx(list('Beta Diversity' = Permanova_NA_G2_vs_neg),
                    '../visualizations/tables/beta_diversity/Beta Diversity Permanova_NA_G2_vs_neg.xlsx')
```
