---
title: "beta_diversity"
author: "Soroush_Torfehnejad"
date: "2026-03-29"
output: html_document
---

# Step 1)

### Load libraries and Data, I rarefied the data based on the conclusions we came to when looking at the QUIME results.The rarefaction number is 5396

```{r Load data and libraries}
library(tidyverse)
library(phyloseq)
library(vegan)
# Load object
ps <- readRDS('../processed_data/phyloseq_taxonomy.rds')


# rarefy the data:
psrare <- ps|>
          rarefy_even_depth(sample.size = 5396, rngseed = 421)
table(sample_sums(psrare))


```

#Step 2) 
### Caclculate the beta diversity using the Bray-Curtis distance matrix.

```{r calculate bray-curtis distances}
# Calculate the distance between each sample pair
ps_WUF <- phyloseq::distance(psrare, method = "wunifrac")
View(as.matrix(ps_WUF)) # Every pairwise comparison included
```

#Step 3) 
### use Multidimensional scaling to reduce the distance matrix into fewer dimentions which will help plot abd understand the Data.

```{r normalize and plot}
# MDS scaling
set.seed(421)
mds = metaMDS(ps_WUF)

# Extract data
mds_data = mds$points %>% as.data.frame %>%
  merge(sample_data(psrare), by='row.names', sort=F)
head(mds_data)
```

#Step 4) 
### Plot the data based on the following comparisons, HA G1, G2 Vs Neg, HA G1 Vs Neg, HA G2 Vs Neg. 

```{r}

# looking at Beta diversity differences between HA group 1, group 2 and negative control

#standardizing all the colors and shapes so they are consistent when comparing graphs. 
custom_colors_HA<-c("1" = "#56B4E9", "2" = "#D55E00", "neg" = "#009E73")
custom_shapes_HA <- c("1" = 21, "2" = 24, "neg" = 22)

# looking at Beta diversity differences between HA group 1, group 2 and uninfected

plot_HA_G1_VS_G2_VS_neg<- mds_data %>%
                                  ggplot(aes(MDS1,MDS2,color = ha_group, fill = ha_group, shape = ha_group )) +
                                  geom_point(size = 1.5, stroke = 1.4) +
                                  scale_color_manual(values = custom_colors_HA)+
                                  scale_fill_manual(values = custom_colors_HA)+
                                  scale_shape_manual(values = custom_shapes_HA)+
                                  stat_ellipse() +
                                  theme_classic(base_size=18)+
                                  theme(axis.title = element_text(size = 12),
                                        axis.text = element_text(size = 10))+
                                  labs( x = "MDS Axis 1",
                                        y = "MDS Axis 2",
                                        title = "Beta Diversity All HA Groups",
                                        color = "HA grouping",
                                        fill = "HA grouping",
                                        shape = "HA grouping")
                           
plot_HA_G1_VS_G2_VS_neg            


# looking at Beta diversity differences between HA group 1 and group 2
plot_HA_G1_VS_G2<- mds_data %>%
                            filter(ha_group != "neg") %>% 
                            ggplot(aes(MDS1,MDS2,color = ha_group, fill = ha_group, shape = ha_group )) +
                            geom_point(size = 1.5, stroke = 1.4) +
                            scale_color_manual(values = custom_colors_HA)+
                            scale_fill_manual(values = custom_colors_HA)+
                            scale_shape_manual(values = custom_shapes_HA)+
                            stat_ellipse() +
                            theme_classic(base_size=18)+
                            theme(axis.title = element_text(size = 12),
                                  axis.text = element_text(size = 10))+
                            labs( x = "MDS Axis 1",
                                  y = "MDS Axis 2",
                                  title = "Beta Diversity HA Group 1 vs Group 2",
                                  color = "HA grouping",
                                  fill = "HA grouping",
                                  shape = "HA grouping")
                  
plot_HA_G1_VS_G2

# looking at Beta diversity differences between HA group 1 and uninfected
plot_HA_G1_VS_NEG<- mds_data %>%
                              filter(ha_group != "2") %>% 
                              ggplot(aes(MDS1,MDS2,color = ha_group, fill = ha_group, shape = ha_group )) +
                              geom_point(size = 1.5, stroke = 1.4) +
                              scale_color_manual(values = custom_colors_HA)+
                              scale_fill_manual(values = custom_colors_HA)+
                              scale_shape_manual(values = custom_shapes_HA)+
                              stat_ellipse() +
                              theme_classic(base_size=18)+
                              theme(axis.title = element_text(size = 12),
                                    axis.text = element_text(size = 10))+
                              labs( x = "MDS Axis 1",
                                    y = "MDS Axis 2",
                                    title = "Beta Diversity HA Group 1 Vs Uninfected",
                                    color = "HA grouping",
                                    fill = "HA grouping",
                                    shape = "HA grouping")

plot_HA_G1_VS_NEG
# looking at Beta diversity differences between HA group 2 and uninfected
plot_HA_G2_VS_NEG<- mds_data %>%
                              filter(ha_group != "1") %>% 
                              ggplot(aes(MDS1,MDS2,color = ha_group, fill = ha_group, shape = ha_group )) +
                              geom_point(size = 1.5, stroke = 1.4) +
                              scale_color_manual(values = custom_colors_HA)+
                              scale_fill_manual(values = custom_colors_HA)+
                              scale_shape_manual(values = custom_shapes_HA)+
                              stat_ellipse() +
                              theme_classic(base_size=18)+
                              theme(axis.title = element_text(size = 12),
                                    axis.text = element_text(size = 10))+
                              labs( x = "MDS Axis 1",
                                    y = "MDS Axis 2",
                                    title = "Beta Diversity HA Group 2 Vs Uninfected",
                                    color = "HA grouping",
                                    fill = "HA grouping",
                                    shape = "HA grouping")
plot_HA_G2_VS_NEG
```

#Step 5)

### Perform a PERMANOVA test to find statistical significance between the HA groups using an Adonis function.

```{r}
metadata <- sample_data(psrare) %>% data.frame() # to make the next lines easier


set.seed(421)
# Single variable anova 

Permanova_on_HA_groups <- adonis2(ps_WUF ~ ha_group, data = metadata, permutations = 99999)

Permanova_on_HA_groups

# when observing if their is a difference in the Beta diversity between HA Group 1,2 or negative control statistical significance was seen p= 0.02326


#this statistical significance led to us comparing each individual group within the ha_group column: group1 vs group 2 vs uninfected


# Comparing HA G1 Vs G2 using PERMANOVA 
compare_G1_vs_G2_HA <- c('2','1')

# Correct the data to make them have the same size when comparing G1 Vs G2
keep_G1_vs_G2_HA <- metadata %>% filter(ha_group %in% compare_G1_vs_G2_HA) %>% rownames()


ps_WUF_G1_vs_G2_HA <- as.matrix(ps_WUF)[keep_G1_vs_G2_HA,keep_G1_vs_G2_HA] %>% as.dist()

# Run a PERMANOVA on G1 vs G2
Permanova_HA_G1_vs_G2 <- adonis2(ps_WUF_G1_vs_G2_HA ~ ha_group,
                       data <- metadata %>% filter(ha_group %in% compare_G1_vs_G2_HA),permutations = 99999)

Permanova_HA_G1_vs_G2 

# This one showed no significance p= 0.3286 indcation no differences between the beta diversity of G1 vs G2

# Comparing HA G1 Vs Neg using PERMANOVA 
compare_G1_vs_neg_HA <- c('neg','1')

# Correct the data to make them have the same size when comparing G1 Vs neg
keep_G1_vs_neg_HA <- metadata %>% filter(ha_group %in% compare_G1_vs_neg_HA) %>% rownames()


ps_WUF_G1_vs_neg_HA <- as.matrix(ps_WUF)[keep_G1_vs_neg_HA,keep_G1_vs_neg_HA] %>% as.dist()

# Run a PERMANOVA on G1 vs neg
Permanova_HA_G1_vs_neg <- adonis2(ps_WUF_G1_vs_neg_HA ~ ha_group,
                       data = metadata %>% filter(ha_group %in% compare_G1_vs_neg_HA), permutations = 99999)

Permanova_HA_G1_vs_neg 


# This one showed no significance p= 0.09864 indication no differences between the beta diversity of G1 vs neg


# Comparing HA G2 Vs Neg using PERMANOVA 
compare_G2_vs_neg_HA <- c('neg','2')

# Correct the data to make them have the same size when comparing G2 Vs neg
keep_G2_vs_neg_HA <- metadata %>% filter(ha_group %in% compare_G2_vs_neg_HA) %>% rownames()


ps_WUF_G2_vs_neg_HA <- as.matrix(ps_WUF)[keep_G2_vs_neg_HA,keep_G2_vs_neg_HA] %>% as.dist()

# Run a PERMANOVA on G2 vs neg
Permanova_HA_G2_vs_neg <- adonis2(ps_WUF_G2_vs_neg_HA ~ ha_group,
                       data = metadata %>% filter(ha_group %in% compare_G2_vs_neg_HA), permutations = 99999)

Permanova_HA_G2_vs_neg 

# This one showed  significance p= 0.0142 indication a statisticaly significant differences between the beta diversity of G2 vs neg. This indicates that group 2 HA influenza viruses may have a stronger affect in changing the beta diversity of the coacle microbiome. 


```




#Step 6
###Plot the data based on the following comparisons, NA G1, G2 Vs Neg, NA G1 Vs Neg, NA G2 Vs Neg. 
```{r}



# looking at Beta diversity differences between NA group 1, group 2 and negative control

#standardizing all the colors and shapes so they are consistent when comparing graphs. 
custom_colors_NA<-c("1" = "darkgrey", "2" = "#CC79A7", "neg" = "#009E73")
custom_shapes_NA <- c("1" = 21, "2" = 24, "neg" = 22)

# looking at Beta diversity differences between HA group 1, group 2 and uninfected

plot_NA_G1_VS_G2_VS_neg<- mds_data %>%
                                  ggplot(aes(MDS1,MDS2,color = na_group, fill = na_group, shape = na_group )) +
                                  geom_point(size = 1.5, stroke = 1.4) +
                                  scale_color_manual(values = custom_colors_NA)+
                                  scale_fill_manual(values = custom_colors_NA)+
                                  scale_shape_manual(values = custom_shapes_NA)+
                                  stat_ellipse() +
                                  theme_classic(base_size=18)+
                                  theme(axis.title = element_text(size = 12),
                                        axis.text = element_text(size = 10))+
                                  labs( x = "MDS Axis 1",
                                        y = "MDS Axis 2",
                                        title = "Beta Diversity All NA Groups",
                                        color = "NA grouping",
                                        fill = "NA grouping",
                                        shape = "NA grouping")
                           
plot_NA_G1_VS_G2_VS_neg            


# looking at Beta diversity differences between HA group 1 and group 2
plot_NA_G1_VS_G2<- mds_data %>%
                            filter(na_group != "neg") %>% 
                            ggplot(aes(MDS1,MDS2,color = na_group, fill = na_group, shape = na_group )) +
                            geom_point(size = 1.5, stroke = 1.4) +
                            scale_color_manual(values = custom_colors_NA)+
                            scale_fill_manual(values = custom_colors_NA)+
                            scale_shape_manual(values = custom_shapes_NA)+
                            stat_ellipse() +
                            theme_classic(base_size=18)+
                            theme(axis.title = element_text(size = 12),
                                  axis.text = element_text(size = 10))+
                            labs( x = "MDS Axis 1",
                                  y = "MDS Axis 2",
                                  title = "Beta Diversity NA Group 1 vs Group 2",
                                  color = "NA grouping",
                                  fill = "NA grouping",
                                  shape = "NA grouping")
                  
plot_NA_G1_VS_G2

# looking at Beta diversity differences between HA group 1 and uninfected
plot_NA_G1_VS_NEG<- mds_data %>%
                              filter(na_group != "2") %>% 
                              ggplot(aes(MDS1,MDS2,color = na_group, fill = na_group, shape = na_group )) +
                              geom_point(size = 1.5, stroke = 1.4) +
                              scale_color_manual(values = custom_colors_NA)+
                              scale_fill_manual(values = custom_colors_NA)+
                              scale_shape_manual(values = custom_shapes_NA)+
                              stat_ellipse() +
                              theme_classic(base_size=18)+
                              theme(axis.title = element_text(size = 12),
                                    axis.text = element_text(size = 10))+
                              labs( x = "MDS Axis 1",
                                    y = "MDS Axis 2",
                                    title = "Beta Diversity NA G1 Vs Uninfected",
                                    color = "NA grouping",
                                    fill = "NA grouping",
                                    shape = "NA grouping")

plot_NA_G1_VS_NEG
# looking at Beta diversity differences between HA group 2 and uninfected
plot_NA_G2_VS_NEG<- mds_data %>%
                              filter(na_group != "1") %>% 
                              ggplot(aes(MDS1,MDS2,color = na_group, fill = na_group, shape = na_group )) +
                              geom_point(size = 1.5, stroke = 1.4) +
                              scale_color_manual(values = custom_colors_NA)+
                              scale_fill_manual(values = custom_colors_NA)+
                              scale_shape_manual(values = custom_shapes_NA)+
                              stat_ellipse() +
                              theme_classic(base_size=18)+
                              theme(axis.title = element_text(size = 12),
                                    axis.text = element_text(size = 10))+
                              labs( x = "MDS Axis 1",
                                    y = "MDS Axis 2",
                                    title = "Beta Diversity NA Group 2 Vs Uninfected",
                                    color = "NA grouping",
                                    fill = "NA grouping",
                                    shape = "NA grouping")
plot_NA_G2_VS_NEG



```
#Step 7) 
##stats for NA group
```{r}

metadata <- sample_data(psrare) %>% data.frame() # to make the next lines easier


set.seed(421)
# Single variable anova 

Permanova_on_NA_groups <- adonis2(ps_WUF ~ na_group, data = metadata, permutation = 99999)

Permanova_on_NA_groups

# when observing if their is a difference in the Beta diversity between NA Group 1,2 or negative control statistical significance was seen p= 0.0235


#this statistical significance led to us comparing each individual group within the na_group column: group1 vs group 2 vs uninfected


# Comparing NA G1 Vs G2 using PERMANOVA 
compare_G1_vs_G2_NA <- c('2','1')

# Correct the data to make them have the same size when comparing G1 Vs G2
keep_G1_vs_G2_NA <- metadata %>% filter(na_group %in% compare_G1_vs_G2_NA) %>% rownames()


ps_WUF_G1_vs_G2_NA <- as.matrix(ps_WUF)[keep_G1_vs_G2_NA,keep_G1_vs_G2_NA] %>% as.dist()

# Run a PERMANOVA on G1 vs G2
Permanova_NA_G1_vs_G2 <- adonis2(ps_WUF_G1_vs_G2_NA ~ na_group,
                       data = metadata %>% filter(na_group %in% compare_G1_vs_G2_NA), permutation = 99999)

Permanova_NA_G1_vs_G2 

# This one showed no significance p= 0.2812 indcation no differences between the beta diversity of G1 vs G2

# Comparing HA G1 Vs Neg using PERMANOVA 
compare_G1_vs_neg_NA <- c('neg','1')

# Correct the data to make them have the same size when comparing G1 Vs neg
keep_G1_vs_neg_NA <- metadata %>% filter(ha_group %in% compare_G1_vs_neg_NA) %>% rownames()


ps_WUF_G1_vs_neg_NA <- as.matrix(ps_WUF)[keep_G1_vs_neg_NA,keep_G1_vs_neg_NA] %>% as.dist()

# Run a PERMANOVA on G1 vs neg
Permanova_NA_G1_vs_neg <- adonis2(ps_WUF_G1_vs_neg_NA ~ na_group,
                       data = metadata %>% filter(ha_group %in% compare_G1_vs_neg_NA), permutations = 99999)

Permanova_NA_G1_vs_neg 


# This one showed no significance p= 0.1906 indcation no differences between the beta diversity of G1 vs neg


# Comparing HA G2 Vs Neg using PERMANOVA 
compare_G2_vs_neg_NA <- c('neg','2')

# Correct the data to make them have the same size when comparing G2 Vs neg
keep_G2_vs_neg_NA <- metadata %>% filter(ha_group %in% compare_G2_vs_neg_NA) %>% rownames()


ps_WUF_G2_vs_neg_NA <- as.matrix(ps_WUF)[keep_G2_vs_neg_NA,keep_G2_vs_neg_NA] %>% as.dist()

# Run a PERMANOVA on G2 vs neg
Permanova_NA_G2_vs_neg <- adonis2(ps_WUF_G2_vs_neg_NA ~ na_group,
                       data = metadata %>% filter(ha_group %in% compare_G2_vs_neg_NA), permutations = 99999)

Permanova_NA_G2_vs_neg 

# This one showed significance p= 0.03712 indcation a statisticaly significant differences between the beta diversity of G2 vs neg. This indicates that group 2 HA influenza viruses may have a stronger affect in changing the beta diversity of the coacle microbiome. 





```



# Step 8)
##save data

```{r}
# saving the data 

#  Save plot

#HA data
ggsave('../visualizations/figures/beta diversity/Beta_Diversity_Graphs- Beta Diversity HA G1 vs G2 vs NEG.jpeg',
       plot = plot_HA_G1_VS_G2_VS_neg,
       height= 4, width = 6.5)

ggsave('../visualizations/figures/beta diversity/Beta_Diversity_Graphs- Beta Diversity HA G1 vs G2.jpeg',
       plot = plot_HA_G1_VS_G2,
       height= 4, width = 6.5)


ggsave('../visualizations/figures/beta diversity/Beta_Diversity_Graphs- HA G1 vs NEG.jpeg',
       plot = plot_HA_G1_VS_NEG,
       height= 4, width = 6.5)


ggsave('../visualizations/figures/beta diversity/Beta_Diversity_Graphs- HA G2 vs NEG.jpeg',
       plot = plot_HA_G2_VS_NEG,
       height= 4, width = 6.5)


#NA data
ggsave('../visualizations/figures/beta diversity/Beta_Diversity_Graphs- Beta Diversity NA G1 vs G2 vs NEG.jpeg',
       plot = plot_NA_G1_VS_G2_VS_neg,
       height= 4, width = 6.5)

ggsave('../visualizations/figures/beta diversity/Beta_Diversity_Graphs- Beta Diversity NA G1 vs G2.jpeg',
       plot = plot_NA_G1_VS_G2,
       height= 4, width = 6.5)


ggsave('../visualizations/figures/beta diversity/Beta_Diversity_Graphs- NA G1 vs NEG.jpeg',
       plot = plot_NA_G1_VS_NEG,
       height= 4, width = 6.5)


ggsave('../visualizations/figures/beta diversity/Beta_Diversity_Graphs- NA G2 vs NEG.jpeg',
       plot = plot_NA_G2_VS_NEG,
       height= 4, width = 6.5)




# Save statistics as an Excel file
writexl::write_xlsx(list('Beta Diversity' = Permanova_on_HA_groups),
                    '../visualizations/tables/Beta diversity stats/Beta_Diversity_Stats - Beta Diversity Permanova_on_HA_groups.xlsx')

writexl::write_xlsx(list('Beta Diversity' = Permanova_HA_G1_vs_G2),
                    '../visualizations/tables/Beta diversity stats/Beta_Diversity_Stats - Beta Diversity Permanova_HA_G1_vs_G2.xlsx')

writexl::write_xlsx(list('Beta Diversity' = Permanova_HA_G1_vs_neg),
                    '../visualizations/tables/Beta diversity stats/Beta_Diversity_Stats - Beta Diversity Permanova_HA_G1_vs_neg.xlsx')

writexl::write_xlsx(list('Beta Diversity' = Permanova_HA_G2_vs_neg),
                    '../visualizations/tables/Beta diversity stats/Beta_Diversity_Stats - Beta Diversity Permanova_HA_G2_vs_neg.xlsx')


writexl::write_xlsx(list('Beta Diversity' = Permanova_on_NA_groups),
                    '../visualizations/tables/Beta diversity stats/Beta_Diversity_Stats - Beta Diversity Permanova_on_NA_groups.xlsx')

writexl::write_xlsx(list('Beta Diversity' = Permanova_NA_G1_vs_G2),
                    '../visualizations/tables/Beta diversity stats/Beta_Diversity_Stats - Beta Diversity Permanova_NA_G1_vs_G2.xlsx')

writexl::write_xlsx(list('Beta Diversity' = Permanova_NA_G1_vs_neg),
                    '../visualizations/tables/Beta diversity stats/Beta_Diversity_Stats - Beta Diversity Permanova_NA_G1_vs_neg.xlsx')

writexl::write_xlsx(list('Beta Diversity' = Permanova_NA_G2_vs_neg),
                    '../visualizations/tables/Beta diversity stats/Beta_Diversity_Stats - Beta Diversity Permanova_NA_G2_vs_neg.xlsx')




```





