---
title: "differential_abundance"
output: html_document
---

# Differential Abundance Analysis using ANCOMBC2

## Loading in Packages

```{r}
library(tidyverse)
library(phyloseq)
library(ANCOMBC)
library(microbiome)
library(ggplot2)
library(ggrepel)
```

Wrangling steps

```{r}
ps = readRDS('../processed_data/phyloseq_taxonomy.rds')
```

```{r}
# setting seed
set.seed(412)


ps_genus = tax_glom(ps, 'Genus')

temp_relab = transform(ps_genus, 'compositional')
temp_filt = filter_taxa(temp_relab, function(x) mean(x) >= 0.0001, TRUE)

ps_filt = prune_taxa(taxa_names(temp_filt),ps_genus)
ps_filt
```

# HA Groups

## Running ANCOMBC2 Analysis

There are three comparisons to be made for HA groups:

-   HA group 1 vs HA group 2

-   HA group 1 vs uninfected (negative)

-   HA group 2 vs uninfected (negative)

ANCOMBC2 on default takes the first alphabetical group in ha_group as the reference, meaning without modifications it will take HA group 1 as the reference. To compare HA group 2 vs uninfected, Frederick Huang (author of ANCOMBC2) suggested subsetting the data.

```{r}
### HA Group 1 as reference ###
ha1_out = ancombc2(data = ps_filt, 
               fix_formula = 'ha_group', 
               p_adj_method = 'BH',
               prv_cut = 0.3) # increased due to small sample size

ha1_statistical_table = ha1_out$res

### HA Group 2 as reference ###
# subset to only group2 and neg
# keep only group2 and neg samples
sub_ps_ha  <- ps_filt %>%
  subset_samples(ha_group %in% c("2", "neg"))

ha2_out = ancombc2(data = sub_ps_ha, 
               fix_formula = 'ha_group', 
               p_adj_method = 'BH',
               prv_cut = 0.3) # increased due to small sample size

ha2_statistical_table = ha2_out$res
```

## Volcano Plots

```{r}
### HA Group1 vs HA Group2 ### 
ha1_statistical_table$association_ = ifelse(ha1_statistical_table$q_ha_group2 < 0.05 & ha1_statistical_table$lfc_ha_group2 > 0, "Enriched", "Not significant")
ha1_statistical_table$association = ifelse(ha1_statistical_table$q_ha_group2 < 0.05 & ha1_statistical_table$lfc_ha_group2 < 0, "Depleted",ha1_statistical_table$association)

vplot_h1h2 <- ggplot(ha1_statistical_table, aes(x = as.numeric(lfc_ha_group2), y = -log10(as.numeric(q_ha_group2)), color = association)) +
  geom_point() +
  labs(x = "Log2 Fold Change", y = "-log10(p-value)",
       title = "Volcano Plot H1 vs H2") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  theme_bw()


vplot_h1h2

### HA Group1 vs neg ### 
ha1_statistical_table$association_ = ifelse(ha1_statistical_table$q_ha_groupneg < 0.05 & ha1_statistical_table$lfc_ha_groupneg > 0, "Enriched", "Not significant")
ha1_statistical_table$association = ifelse(ha1_statistical_table$q_ha_groupneg < 0.05 & ha1_statistical_table$lfc_ha_groupneg < 0, "Depleted",ha1_statistical_table$association)

vplot_h1neg <- ggplot(ha1_statistical_table, aes(x = as.numeric(lfc_ha_groupneg), y = -log10(as.numeric(q_ha_groupneg)), color = association)) +
  geom_point() +
  labs(x = "Log2 Fold Change", y = "-log10(p-value)",
       title = "Volcano Plot H1 vs neg") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  theme_bw()


vplot_h1neg

### HA Group2 vs neg ###
ha2_statistical_table$association_ = ifelse(ha2_statistical_table$q_ha_groupneg < 0.05 & ha2_statistical_table$lfc_ha_groupneg > 0, "Enriched", "Not significant")
ha2_statistical_table$association = ifelse(ha2_statistical_table$q_ha_groupneg < 0.05 & ha2_statistical_table$lfc_ha_groupneg < 0, "Depleted", ha2_statistical_table$association)

vplot_h2neg <- ggplot(ha2_statistical_table, aes(x = as.numeric(lfc_ha_groupneg), y = -log10(as.numeric(q_ha_groupneg)), color = association)) +
  geom_point() +
  labs(x = "Log2 Fold Change", y = "-log10(p-value)",
       title = "Volcano Plot H2 vs neg") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  theme_bw()


vplot_h2neg
```

## Log2Fold Change Plots

```{r}
# extract Genus taxonomy from phyloseq for downstream plots
genus_tax <- data.frame(
  taxon = taxa_names(ps_filt),
  Genus = as.data.frame(tax_table(ps_filt))$Genus
)

### HA Group 1 vs HA Group 2 ### 
# Filter stats to only include taxa that are differentially abundant between H1 group and H2 group
h1h2_taxa_plot = ha1_statistical_table %>% 
  filter(diff_robust_ha_group2==T) %>%
  left_join(genus_tax, by = "taxon")


# Plot the log 2 fold changes

h1h2_taxa_plot %>%
  ggplot(aes(Genus,lfc_ha_group2)) +
  geom_col() +
  coord_flip()

### HA Group 1 vs neg ### 
# Filter stats to only include taxa that are differentially abundant between H1 group and negative group
h1neg_taxa_plot = ha1_statistical_table %>% 
  filter(diff_robust_ha_groupneg==T) %>%
  left_join(genus_tax, by = "taxon")

# Plot the log 2 fold changes

h1neg_taxa_plot %>%
  ggplot(aes(Genus,lfc_ha_groupneg)) +
  geom_col() +
  coord_flip()


### HA Group 2 vs neg ### 
# Filter stats to only include taxa that are differentially abundant between H2 group and negative group
h2neg_taxa_plot = ha2_statistical_table %>% 
  filter(diff_robust_ha_groupneg==T) %>%
  left_join(genus_tax, by = "taxon")

# Plot the log 2 fold changes

h2neg_taxa_plot %>%
  ggplot(aes(Genus,lfc_ha_groupneg)) +
  geom_col() +
  coord_flip()
```

Improved Plots

```{r}
### HA Group 1 vs Neg ###
df_h1neg = ha1_statistical_table %>%
  left_join(genus_tax, by = "taxon") 

df_fig_h1neg = df_h1neg %>%
    filter(diff_robust_ha_groupneg == T) %>% 
    mutate(direct = ifelse(lfc_ha_groupneg > 0, "Positive LFC", "Negative LFC"),
                  color = ifelse(passed_ss_ha_groupneg == 1, "aquamarine3", "black"),
                  Genus = reorder(Genus, -lfc_ha_groupneg))
df_fig_h1neg$direct = factor(df_fig_h1neg$direct, 
                           levels = c("Positive LFC", "Negative LFC"))
  
fig_h1neg = df_fig_h1neg %>%
    ggplot(aes(x = Genus, y = lfc_ha_groupneg, fill = direct)) + 
    geom_bar(stat = "identity", width = 0.7, color = "black", 
             position = position_dodge(width = 0.4)) +
    geom_errorbar(aes(ymin = lfc_ha_groupneg - se_ha_groupneg, ymax = lfc_ha_groupneg + se_ha_groupneg), 
                  width = 0.2, position = position_dodge(0.05), color = "black") + 
    labs(x = NULL, y = "Log2 Fold Change", 
         title = "Log2 FC HA Group 1/Uninfected") + 
    scale_fill_discrete(name = NULL) +
    scale_color_discrete(name = NULL) +
    theme_bw() + 
    theme(plot.title = element_text(hjust = 0.5, size = 30),
          panel.grid.minor.y = element_blank(),
          axis.text.x = element_text(angle = 60, hjust = 1, size = 18),
          axis.title.y = element_text(size = 18),
          axis.text.y = element_text(hjust=1, size = 18),
          legend.text = element_text(hjust = 1, size = 12))
fig_h1neg

### HA Group 2 vs Neg ###
df_h2neg = ha2_statistical_table %>%
  left_join(genus_tax, by = "taxon") 

df_fig_h2neg = df_h2neg %>%
    filter(diff_robust_ha_groupneg == T) %>% 
    mutate(direct = ifelse(lfc_ha_groupneg > 0, "Positive LFC", "Negative LFC"),
                  color = ifelse(passed_ss_ha_groupneg == 1, "aquamarine3", "black"),
                  Genus = reorder(Genus, -lfc_ha_groupneg))

df_fig_h2neg$direct = factor(df_fig_h2neg$direct, 
                           levels = c("Positive LFC", "Negative LFC"))
  
fig_h2neg = df_fig_h2neg %>%
    ggplot(aes(x = Genus, y = lfc_ha_groupneg, fill = direct)) + 
    geom_bar(stat = "identity", width = 0.7, color = "black", 
             position = position_dodge(width = 0.4)) +
    geom_errorbar(aes(ymin = lfc_ha_groupneg - se_ha_groupneg, ymax = lfc_ha_groupneg + se_ha_groupneg), 
                  width = 0.2, position = position_dodge(0.05), color = "black") + 
    labs(x = NULL, y = "Log2 Fold Change", 
         title = "Log2 FC HA Group 2/Uninfected") + 
    scale_fill_discrete(name = NULL) +
    scale_color_discrete(name = NULL) +
    theme_bw() + 
    theme(plot.title = element_text(hjust = 0.5, size = 30),
          panel.grid.minor.y = element_blank(),
          axis.text.x = element_text(angle = 60, hjust = 1, size = 18),
          axis.title.y = element_text(size = 18),
          axis.text.y = element_text(hjust=1, size = 18),
          legend.text = element_text(hjust = 1, size = 12))
fig_h2neg
```

# NA Groups

## Running ANCOMBC2 Analysis

There are three comparisons to be made for HA groups:

-   NA group 1 vs NA group 2

-   NA group 1 vs uninfected (negative)

-   NA group 2 vs uninfected (negative)

```{r}
### NA Group 1 as reference ###
na1_out = ancombc2(data = ps_filt, 
               fix_formula = 'na_group', 
               p_adj_method = 'BH',
               prv_cut = 0.3) # increased due to small sample size

na1_statistical_table = na1_out$res

### NA Group 2 as reference ###
sub_ps_na <- subset_samples(ps_filt, na_group %in% c("2", "neg")) # subseting data to remove ha group_1

na2_out = ancombc2(data = sub_ps_na, 
               fix_formula = 'na_group', 
               p_adj_method = 'BH',
               prv_cut = 0.3) # increased due to small sample size

na2_statistical_table = na2_out$res
```

## Volcano Plots

```{r}
### NA Group1 vs NA Group2 ### 
na1_statistical_table$association_ = ifelse(na1_statistical_table$q_na_group2 < 0.05 & na1_statistical_table$lfc_na_group2 > 0, "Enriched", "Not significant")
na1_statistical_table$association = ifelse(na1_statistical_table$q_na_group2 < 0.05 & na1_statistical_table$lfc_na_group2 < 0, "Depleted",na1_statistical_table$association)

vplot_n1n2 <- ggplot(na1_statistical_table, aes(x = as.numeric(lfc_na_group2), y = -log10(as.numeric(q_na_group2)), color = association)) +
  geom_point() +
  labs(x = "Log2 Fold Change", y = "-log10(p-value)",
       title = "Volcano Plot N1 vs N2") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  theme_bw()


vplot_n1n2

### NA Group1 vs neg ### 
na1_statistical_table$association_ = ifelse(na1_statistical_table$q_na_groupneg < 0.05 & na1_statistical_table$lfc_na_groupneg > 0, "Enriched", "Not significant")
na1_statistical_table$association = ifelse(na1_statistical_table$q_na_groupneg < 0.05 & na1_statistical_table$lfc_na_groupneg < 0, "Depleted",na1_statistical_table$association)

vplot_n1neg <- ggplot(na1_statistical_table, aes(x = as.numeric(lfc_na_groupneg), y = -log10(as.numeric(q_na_groupneg)), color = association)) +
  geom_point() +
  labs(x = "Log2 Fold Change", y = "-log10(p-value)",
       title = "Volcano Plot N1 vs neg") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  theme_bw()


vplot_n1neg

### NA Group2 vs neg ###
na2_statistical_table$association_ = ifelse(na2_statistical_table$q_na_groupneg < 0.05 & na2_statistical_table$lfc_na_groupneg > 0, "Enriched", "Not significant")
na2_statistical_table$association = ifelse(na2_statistical_table$q_na_groupneg < 0.05 & na2_statistical_table$lfc_na_groupneg < 0, "Depleted", na2_statistical_table$association)

vplot_h2neg <- ggplot(na2_statistical_table, aes(x = as.numeric(lfc_na_groupneg), y = -log10(as.numeric(q_na_groupneg)), color = association)) +
  geom_point() +
  labs(x = "Log2 Fold Change", y = "-log10(p-value)",
       title = "Volcano Plot N2 vs neg") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  theme_bw()


vplot_h2neg
```

## Log2Fold Change Plots

```{r}
### NA Group 1 vs NA Group 2 ### 
# Filter stats to only include taxa that are differentially abundant between N1 group and N2 group
n1n2_taxa_plot = na1_statistical_table %>% 
  filter(diff_robust_na_group2==T) %>%
  left_join(genus_tax, by = "taxon")

# Plot the log 2 fold changes

n1n2_taxa_plot %>%
  ggplot(aes(Genus,lfc_na_group2)) +
  geom_col() +
  coord_flip()

### NA Group 1 vs neg ### 
# Filter stats to only include taxa that are differentially abundant between N1 group and negative group
n1neg_taxa_plot = na1_statistical_table %>% 
  filter(diff_robust_na_groupneg==T) %>%
  left_join(genus_tax, by = "taxon")

# Plot the log 2 fold changes

n1neg_taxa_plot %>%
  ggplot(aes(Genus,lfc_na_groupneg)) +
  geom_col() +
  coord_flip()


### NA Group 2 vs neg ### 
# Filter stats to only include taxa that are differentially abundant between N2 group and negative group
n2neg_taxa_plot = na2_statistical_table %>% 
  filter(diff_robust_na_groupneg==T) %>%
  left_join(genus_tax, by = "taxon")

# Plot the log 2 fold change #

n2neg_taxa_plot %>%
  ggplot(aes(x=Genus, y=lfc_na_groupneg)) +
  geom_col() +
  coord_flip()
```

Improved Plots

```{r}
### NA Group 1 vs Neg ###
df_n1neg = na1_statistical_table %>%
  left_join(genus_tax, by = "taxon") 

df_fig_n1neg = df_n1neg %>%
    filter(diff_robust_na_groupneg == T) %>% 
    mutate(direct = ifelse(lfc_na_groupneg > 0, "Positive LFC", "Negative LFC"),
                  color = ifelse(passed_ss_na_groupneg == 1, "aquamarine3", "black"),
                  Genus = reorder(Genus, -lfc_na_groupneg))
df_fig_n1neg$direct = factor(df_fig_n1neg$direct, 
                           levels = c("Positive LFC", "Negative LFC"))
  
fig_n1neg = df_fig_n1neg %>%
    ggplot(aes(x = Genus, y = lfc_na_groupneg, fill = direct)) + 
    geom_bar(stat = "identity", width = 0.7, color = "black", 
             position = position_dodge(width = 0.4)) +
    geom_errorbar(aes(ymin = lfc_na_groupneg - se_na_groupneg, ymax = lfc_na_groupneg + se_na_groupneg), 
                  width = 0.2, position = position_dodge(0.05), color = "black") + 
    labs(x = NULL, y = "Log2 Fold Change", 
         title = "Log2 FC NA Group 1/Uninfected") + 
    scale_fill_discrete(name = NULL) +
    scale_color_discrete(name = NULL) +
    theme_bw() + 
        theme(plot.title = element_text(hjust = 0.5, size = 30),
          panel.grid.minor.y = element_blank(),
          axis.text.x = element_text(angle = 60, hjust = 1, size = 18),
          axis.title.y = element_text(size = 18),
          axis.text.y = element_text(hjust=1, size = 18),
          legend.text = element_text(hjust = 1, size = 12))
fig_n1neg

### NA Group 2 vs Neg ###
df_n2neg = na2_statistical_table %>%
  left_join(genus_tax, by = "taxon") 

df_fig_n2neg = df_n2neg %>%
    filter(diff_robust_na_groupneg == T) %>% 
    mutate(direct = ifelse(lfc_na_groupneg > 0, "Positive LFC", "Negative LFC"),
                  color = ifelse(passed_ss_na_groupneg == 1, "aquamarine3", "black"),
                  Genus = reorder(Genus, -lfc_na_groupneg))

df_fig_n2neg$direct = factor(df_fig_n2neg$direct, 
                           levels = c("Positive LFC", "Negative LFC"))
  
fig_n2neg = df_fig_n2neg %>%
    ggplot(aes(x = Genus, y = lfc_na_groupneg, fill = direct)) + 
    geom_bar(stat = "identity", width = 0.7, color = "black", 
             position = position_dodge(width = 0.4)) +
    geom_errorbar(aes(ymin = lfc_na_groupneg - se_na_groupneg, ymax = lfc_na_groupneg + se_na_groupneg), 
                  width = 0.2, position = position_dodge(0.05), color = "black") + 
    labs(x = NULL, y = "Log2 Fold Change", 
         title = "Log2 FC NA Group 2/Uninfected") + 
    scale_fill_discrete(name = NULL) +
    scale_color_discrete(name = NULL) +
    theme_bw() + 
        theme(plot.title = element_text(hjust = 0.5, size = 30),
          panel.grid.minor.y = element_blank(),
          axis.text.x = element_text(angle = 60, hjust = 1, size = 18),
          axis.title.y = element_text(size = 18),
          axis.text.y = element_text(hjust=1, size = 18),
          legend.text = element_text(hjust = 1, size = 12))
fig_n2neg
```

Saving Graphs Generated

```{r}
ggsave("../visualizations/figures/differential_abundance_h1vsneg.png", plot = fig_h1neg, width = 18, height = 10)
ggsave("../visualizations/figures/differential_abundance_h2vsneg.png", plot = fig_h2neg, width = 18, height = 10)
ggsave("../visualizations/figures/differential_abundance_n1vsneg.png", plot = fig_n1neg, width = 18, height = 10)
ggsave("../visualizations/figures/differential_abundance_n2vsneg.png", plot = fig_n2neg, width = 18, height = 10)


```
