---
title: "differential_abundance"
output: html_document
---

# Differential Abundance Analysis using ANCOMBC2

## Loading in Packages

```{r}
library(tidyverse)
library(phyloseq)
library(ANCOMBC)
library(microbiome)
library(ggplot2)
library(ggrepel)
```

Wrangling steps

```{r}
ps = readRDS('../processed_data/phyloseq_taxonomy.rds')
```

```{r}
# setting seed
set.seed(412)


ps_genus = tax_glom(ps, 'Genus')

temp_relab = transform(ps_genus, 'compositional')
temp_filt = filter_taxa(temp_relab, function(x) mean(x) >= 0.0001, TRUE)

ps_filt = prune_taxa(taxa_names(temp_filt),ps_genus)
ps_filt
```

# HA Groups

## Running ANCOMBC2 Analysis

There are three comparisons to be made for HA groups:

-   HA group 1 vs HA group 2

-   HA group 1 vs uninfected (negative)

-   HA group 2 vs uninfected (negative)

ANCOMBC2 on default takes the first alphabetical group in ha_group as the reference, meaning without modifications it will take HA group 1 as the reference. To compare HA group 2 vs uninfected, Frederick Huang (author of ANCOMBC2) suggested subseting the data.

```{r}
### H1 as reference ###
ha1_out = ancombc2(data = ps_filt, 
               fix_formula = 'ha_group', 
               p_adj_method = 'BH',
               prv_cut = 0.3) # increased due to small sample size

ha1_statistical_table = ha1_out$res
view(ha1_statistical_table)

### H2 as reference ###
sub_ps <- subset_samples(ps_filt, ha_group %in% c("group2", "neg")) # subseting data to remove ha group_1

ha2_out = ancombc2(data = ps_filt, 
               fix_formula = 'ha_group', 
               p_adj_method = 'BH',
               prv_cut = 0.3) # increased due to small sample size

ha2_statistical_table = ha2_out$res
view(ha2_statistical_table)
```

## Volcano Plots

```{r}
### H1 vs H2 ### 
ha1_statistical_table$association_ = ifelse(ha1_statistical_table$q_ha_group2 < 0.05 & ha1_statistical_table$lfc_ha_group2 > 0, "Enriched", "Not significant")
ha1_statistical_table$association = ifelse(ha1_statistical_table$q_ha_group2 < 0.05 & ha1_statistical_table$lfc_ha_group2 < 0, "Depleted",ha1_statistical_table$association)

vplot_h1h2 <- ggplot(ha1_statistical_table, aes(x = as.numeric(lfc_ha_group2), y = -log10(as.numeric(q_ha_group2)), color = association)) +
  geom_point() +
  labs(x = "Log2 Fold Change", y = "-log10(p-value)",
       title = "Volcano Plot H1 vs H2") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  theme_bw()


vplot_h1h2

### H1 vs neg ### 
ha1_statistical_table$association_ = ifelse(ha1_statistical_table$q_ha_groupneg < 0.05 & ha1_statistical_table$lfc_ha_groupneg > 0, "Enriched", "Not significant")
ha1_statistical_table$association = ifelse(ha1_statistical_table$q_ha_groupneg < 0.05 & ha1_statistical_table$lfc_ha_groupneg < 0, "Depleted",ha1_statistical_table$association)

vplot_h1neg <- ggplot(ha1_statistical_table, aes(x = as.numeric(lfc_ha_groupneg), y = -log10(as.numeric(q_ha_groupneg)), color = association)) +
  geom_point() +
  labs(x = "Log2 Fold Change", y = "-log10(p-value)",
       title = "Volcano Plot H1 vs neg") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  theme_bw()


vplot_h1neg

### H2 vs neg ###
ha2_statistical_table$association_ = ifelse(ha2_statistical_table$q_ha_groupneg < 0.05 & ha2_statistical_table$lfc_ha_groupneg > 0, "Enriched", "Not significant")
ha2_statistical_table$association = ifelse(ha2_statistical_table$q_ha_groupneg < 0.05 & ha2_statistical_table$lfc_ha_groupneg < 0, "Depleted", ha2_statistical_table$association)

vplot_h2neg <- ggplot(ha2_statistical_table, aes(x = as.numeric(lfc_ha_groupneg), y = -log10(as.numeric(q_ha_groupneg)), color = association)) +
  geom_point() +
  labs(x = "Log2 Fold Change", y = "-log10(p-value)",
       title = "Volcano Plot H2 vs neg") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  theme_bw()


vplot_h2neg
```

## Log2Change Plots

```{r}
### H1 vs neg ### 
# Filter stats to only include taxa that are differentially abundant between the gut and tongue
h1neg_taxa_plot = ha1_statistical_table %>% 
  filter(diff_robust_ha_groupneg==T) 

# Plot the log 2 fold changes
# Note: you'll have to figure out how to plot the genera names instead of the
# representative ASV names!

h1neg_taxa_plot %>%
  ggplot(aes(taxon,lfc_ha_groupneg)) +
  geom_col() +
  coord_flip()


### H2 vs neg ### 
# Filter stats to only include taxa that are differentially abundant between the gut and tongue
h2neg_taxa_plot = ha2_statistical_table %>% 
  filter(diff_robust_ha_groupneg==T) 

# Plot the log 2 fold changes
# Note: you'll have to figure out how to plot the genera names instead of the
# representative ASV names!

h2neg_taxa_plot %>%
  ggplot(aes(taxon,lfc_ha_groupneg)) +
  geom_col() +
  coord_flip()

```

# NA Groups

## Running ANCOMBC2 Analysis

There are three comparisons to be made for HA groups:

-   NA group 1 vs NA group 2

-   NA group 1 vs uninfected (negative)

-   NA group 2 vs uninfected (negative)

```{r}
### H1 as reference ###
na1_out = ancombc2(data = ps_filt, 
               fix_formula = 'na_group', 
               p_adj_method = 'BH',
               prv_cut = 0.3) # increased due to small sample size

na1_statistical_table = na1_out$res
view(na1_statistical_table)

### H2 as reference ###
sub_ps <- subset_samples(ps_filt, na_group %in% c("group2", "neg")) # subseting data to remove ha group_1

na2_out = ancombc2(data = ps_filt, 
               fix_formula = 'na_group', 
               p_adj_method = 'BH',
               prv_cut = 0.3) # increased due to small sample size

na2_statistical_table = na2_out$res
view(na2_statistical_table)
```
