---
title: "functional_potential"
output: html_document
creator: "Alex Elbaz"
---

# Predicted Functional Potential Analyses (PICRUSt2)

## 1. Load Libraries

```{r}
library(tidyverse)
library(phyloseq)
library(ggpicrust2)
library(devtools)
install_github("biobakery/maaslin2")

# A log2 fold change calculation function alongside a modified ggpicrust2 errorbar function, both from the MICB 305 2025W2 GitHub, are used within this script (imports below).
# This code is attributed to the creators of the MICB 305 2025W2 GitHub and is cited within our research paper.
source('reference_functions/ggpicrust2_calculate_log2fc.R')
source('reference_functions/ggpicrust2_errorbar_function_fixed.R')

# Some internal functions appear inaccessible in ggpicrust2, so I am importing the internal functions separately for now
source('reference_functions/data_utils.R')
```

## 2. Load Data

Functional annotations were generated via command-line PICRUSt2 and then exported into this project under the `../processed_data/` directory.

```{r}
metadata = data.frame(readRDS('../processed_data/phyloseq_taxonomy.rds')@sam_data) %>%
  rownames_to_column()

metacyc_path_abun = read.delim('../processed_data/path_abun_unstrat.tsv', row.names = 1)
```

## 3. Set Seed

```{r}
set.seed(1234)
```

## 4. Conduct Functional Potential Analysis (via ggpicrust2)

This analysis examines differences in predicted functional potential between HA group 1 and HA group 2 influenza-infected mallards.

```{r}
# Filter out uninfected samples and rename groups so downstream figure is more interpretable
metadata = filter(metadata, ha_group != "neg")
metadata$ha_group = paste("ha_group", metadata$ha_group, sep="_")

metacyc_path_abun = metacyc_path_abun[, colnames(metacyc_path_abun) %in% metadata$rowname]


# Run pathway differential abundance analysis (Maaslin2 method)
daa_df = pathway_daa(abundance = metacyc_path_abun,
                     metadata = metadata,
                     group = "ha_group",
                     daa_method = "Maaslin2",
                     select = NULL,
                     reference = NULL)

# Get correct log2 fold change values; currently, pathway_daa using the Maaslin2 method uses incorrect log2 fold changes.
log2_fc = fix_lfc(abundance = metacyc_path_abun,
                  metadata = metadata,
                  group = "ha_group",
                  reference = "ha_group_2") %>%
  select(feature, log2_fold_change)

# Add correct log2 fold change values to the differential abundance analysis data frame
daa_df = left_join(daa_df, log2_fc, by="feature")

# Add pathway annotations
daa_annotated_df = pathway_annotation(pathway = "MetaCyc", daa_results_df = daa_df)
```

## 5. Generate Figure (Bar plot with most significant pathways)

```{r}
# Create column containing both pathway names & annotations
daa_annotated_df = daa_annotated_df %>%
  mutate(label = paste(feature, ifelse(is.na(description), "No description available", description), sep = "; "))

# Filter for significant results
daa_annotated_df_filt = filter(daa_annotated_df, abs(log2_fold_change) >= 4)# & p_adjust < 0.05)

# Generate pathway errorbar plot
peb = pathway_errorbar_fixed(abundance = metacyc_path_abun,
                             daa_results_df = daa_annotated_df_filt,
                             Group = metadata$ha_group,
                             order_by_log = TRUE,
                             ko_to_kegg = FALSE,
                             p_value_bar = TRUE,
                             x_lab = "label")
```

## 6. Save Figure

```{r}
ggsave("../visualizations/figures/functional_potential.png", plot = peb, width = 18, height = 10)
```

## 7. Additional Useful Information

```{r}
num_pathways = nrow(daa_annotated_df)
num_signif = nrow(filter(daa_annotated_df, p_adjust < 0.05))
prop_signif = num_signif/num_pathways

print(paste("Pathways analyzed:", num_pathways))
print(paste("Significant pahtways (p_adjust < 0.05):", num_signif))
print(paste("Proportion significant:", round(prop_signif, 2)))
```
