---
title: "functional_potential"
output: html_document
author: "Alex Elbaz"
---

# Predicted Functional Potential Analyses (PICRUSt2)

## 1. Load Libraries

```{r}
library(tidyverse)
library(phyloseq)
library(ggpicrust2)
library(devtools)
install_github("biobakery/maaslin2")

# A log2 fold change calculation function alongside a modified ggpicrust2 errorbar function, both from the MICB 305 2025W2 GitHub, are used within this script (imports below).
# This code is attributed to the creators of the MICB 305 2025W2 GitHub and is cited within our research paper.
source('reference_functions/ggpicrust2_calculate_log2fc.R')
source('reference_functions/ggpicrust2_errorbar_function_fixed.R')
```

## 2. Define Functions

```{r}
# Run entire functional pathway differential abundance pipeline.
# abundance: functional pathway abundance matrix
# metadata: metadata
# group: groups of interest to compare
# method: differential abundance method (default Maaslin2)
# reference: Reference group (for log2 fold change calculations)
pathway_daa_pipeline = function(abundance, metadata, group, method = "Maaslin2", reference) {
  # Run pathway differential abundance analysis
  daa = pathway_daa(abundance = abundance,
                    metadata = metadata,
                     group = group,
                     daa_method = method,
                     select = NULL,
                     reference = NULL) %>%
    select(-log2_fold_change)
  
  # Get correct log2 fold change values; currently, pathway_daa using the Maaslin2 method
  # uses incorrect log2 fold changes.
  log2fc = fix_lfc(abundance = abundance,
                   metadata = metadata,
                   group = group,
                   reference = reference) %>%
    select(feature, log2_fold_change)
  
  # Replace incorrect log2 fold change values with the correct ones
  daa = left_join(daa, log2fc, by="feature") %>%
    relocate(log2_fold_change, .before = adj_method)
  
  # Add pathway annotations & create column containing pathway names and annotations
  daa = pathway_annotation(pathway = "MetaCyc", daa_results_df = daa) %>%
    mutate(label = paste(feature, ifelse(is.na(description),
                                       "No description available",
                                       description),
                       sep = "; "))
  
  return(daa)
}


# Generate pathway error bar paired bar plot formatted appropriately.
# daa: pathway differential abundance results
# abundance: pathway abundance matrix
# metadata_group: metadata data frame column containing groups of interest
# log2fc_thresh: absolute log2 fold change threshold to keep in final figure
# p_adjust_thresh: adjusted p-value threshold to keep in final figure
peb_pipeline = function(daa, abundance, metadata_group, log2fc_thresh, p_adjust_thresh) {
  # Filter for most significant results
  daa_filt = daa %>% filter(abs(log2_fold_change) > log2fc_thresh & p_adjust < p_adjust_thresh)
  
  # Generate pathway error bar plot
  peb = pathway_errorbar_fixed(abundance = abundance,
                               daa_results_df = daa_filt,
                               Group = metadata_group,
                               order_by_log = TRUE,
                               ko_to_kegg = FALSE,
                               p_value_bar = TRUE,
                               p_values_threshold = 1,
                               x_lab = "label")
  return(peb)
}
```

## 3. Load Data & Initial Wrangling

Functional annotations were generated via command-line PICRUSt2 and then exported into this project under the `../processed_data/` directory.

```{r}
# Load metadata; keep all samples
metadata_all = data.frame(readRDS('../processed_data/phyloseq_taxonomy.rds')@sam_data) %>%
  rownames_to_column()

# Rename and order infection status groups so the downstream figures are more interpretable
metadata_all$influenza_status = metadata_all$influenza_status %>%
  replace_values(from = "Pos", to = "infected") %>%
  replace_values(from = "Neg", to = "uninfected") %>%
  factor(levels = c("infected", "uninfected"))

# Make a copy of the metadata; remove uninfected samples
metadata_infected = metadata_all %>% filter(ha_group != "neg" & na_group != "neg")

# Rename grouping variable so the downstream figures are more interpretable
metadata_infected$ha_group = ifelse(metadata_infected$ha_group == "1",
                                    "HA Group 1",
                                    "HA Group 2")
metadata_infected$na_group = ifelse(metadata_infected$na_group == "1",
                                    "NA Group 1",
                                    "NA Group 2")


# Load predicted functional pathway abundances; keep all samples
metacyc_path_abun_all = read.delim('../processed_data/path_abun_unstrat.tsv', row.names = 1)

# Copy predicted functional pathway abundances; remove uninfected samples
metacyc_path_abun_infected = metacyc_path_abun_all[, colnames(metacyc_path_abun_all) %in% metadata_infected$rowname]
```

## 4. Conduct Functional Potential Analysis (via ggpicrust2)

This analysis examines differences in predicted functional potential between uninfected and influenza-infected mallards, HA Group 1 & HA Group 2 influenza-infected mallards, and NA Group 1 & NA Group 2 influenza-infected mallards.

```{r}
# Set seed for reproducible analyses
set.seed(1234)

# Infected vs. Uninfected
daa_infection = pathway_daa_pipeline(metacyc_path_abun_all,
                                     metadata_all,
                                     "influenza_status",
                                     "Maaslin2",
                                     "uninfected")

# HA Group 1 vs. HA Group 2
daa_ha = pathway_daa_pipeline(metacyc_path_abun_infected,
                              metadata_infected,
                              "ha_group",
                              "Maaslin2",
                              "HA Group 2")

# NA Group 1 vs. NA Group 2
daa_na = pathway_daa_pipeline(metacyc_path_abun_infected,
                              metadata_infected,
                              "na_group",
                              "Maaslin2",
                              "NA Group 2")
```

## 5. Generate Primary Figures

Importantly, there were many significant results when comparing infected to uninfected, but **NO** significant results when comparing HA Group 1 to HA Group 2 or when comparing NA Group 1 to NA Group 2. Hence, to produce figures which highlight the most significant pathways while retaining readability, absolute log2 fold change and adjusted p-value cutoffs were set for all analyses:

-   Infected vs. Uninfected: absolute log2 fold change \> 1.85 & adjusted p-value \< 0.05

-   HA Group 1 vs. HA Group 2: absolute log2 fold change \> 2.25 & adjusted p-value \< 0.95

-   NA Group 1 vs. NA Group 2: absolute log2 fold change \> 1.75 & adjusted p-value \< 0.65

```{r}
# Infected vs. Uninfected
peb_infection = peb_pipeline(daa_infection,
                             metacyc_path_abun_all,
                             metadata_all$influenza_status,
                             1.85,
                             0.05)

# HA Group 1 vs. HA Group 2 
peb_ha = peb_pipeline(daa_ha,
                      metacyc_path_abun_infected,
                      metadata_infected$ha_group,
                      2.25,
                      0.95)

# NA Group 1 vs. NA Group 2
peb_na = peb_pipeline(daa_na,
                      metacyc_path_abun_infected,
                      metadata_infected$na_group,
                      1.75,
                      0.65)
```

## 6. Generate Significance & Log2 Fold Change Summary Plots

### 6.1 Adjusted P-value Distributions by Comparison

```{r}
# Aggregate adjusted p-value data into a single data frame
daa_p_adj_summary = data.frame("Uninfected_vs._Infected" = daa_infection$p_adjust,
                               "HA_Group_1_vs._HA_Group_2" = daa_ha$p_adjust,
                               "NA_Group_1_vs._NA_Group_2" = daa_na$p_adjust)

# Pivot aggregated data, repair group names, remove NAs, & order groups as desired for plotting
daa_p_adj_summary = pivot_longer(daa_p_adj_summary,
                                 cols = everything(),
                                 names_to = "group",
                                 values_to = "p_adj") %>%
  na.omit()

daa_p_adj_summary$group = sapply(daa_p_adj_summary$group,
                                 str_replace_all,
                                 pattern = "_",
                                 replacement = " ") %>%
  factor(levels = c("Uninfected vs. Infected",
                    "HA Group 1 vs. HA Group 2",
                    "NA Group 1 vs. NA Group 2"))


# Generate adjusted p-value distribution summary plot
daa_p_adj_summary_plot = ggplot(daa_p_adj_summary, aes(x = p_adj)) +
  geom_histogram(binwidth = 0.02, colour = "black", fill = "lightgrey", linewidth = 0.2) +
  facet_wrap(~group, ncol = 1, scales="free_y") +
  labs(x = "Adjusted P value", y = "Count") +
  theme_bw() +
  theme(text = element_text(size = 14, hjust = 0.5))
```

### 6.2 Log2 Fold Change Distributions by Comparison

```{r}
# Aggregate adjusted p-value data into a single data frame
daa_log2fc_summary = data.frame("Uninfected_vs._Infected" = daa_infection$log2_fold_change,
                                "HA_Group_1_vs._HA_Group_2" = daa_ha$log2_fold_change,
                                "NA_Group_1_vs._NA_Group_2" = daa_na$log2_fold_change)

# Pivot aggregated data, repair group names, remove NAs, & order groups as desired for plotting
daa_log2fc_summary = pivot_longer(daa_log2fc_summary,
                                  cols = everything(),
                                  names_to = "group",
                                  values_to = "log2fc") %>%
  na.omit()

daa_log2fc_summary$group = sapply(daa_log2fc_summary$group,
                                  str_replace_all,
                                  pattern = "_",
                                  replacement = " ") %>%
  factor(levels = c("Uninfected vs. Infected",
                    "HA Group 1 vs. HA Group 2",
                    "NA Group 1 vs. NA Group 2"))


# Generate log2 fold change distribution summary plot
daa_log2fc_summary_plot = ggplot(daa_log2fc_summary, aes(x = log2fc)) +
  geom_histogram(binwidth = 0.25, colour = "black", fill = "lightgrey", linewidth = 0.2) +
  facet_wrap(~group, ncol = 1) +
  labs(x = "Log2 Fold Change", y = "Count") +
  theme_bw() +
  theme(text = element_text(size = 14, hjust = 0.5))
# Mark the means in red on the histograms?
```

## 7. Save Figures

```{r}
# Predicted functional pathway abundance figures
ggsave("../visualizations/figures/functional_analysis/functional_analysis_infected.png", plot = peb_infection, width = 14, height = 8)

ggsave("../visualizations/figures/functional_analysis/functional_analysis_ha.png",
       plot = peb_ha, width = 14, height = 8)

ggsave("../visualizations/figures/functional_analysis/functional_analysis_na.png",
       plot = peb_na, width = 14, height = 8)


# Significance & effect size summary figures
ggsave("../visualizations/figures/functional_analysis/p_adj_distribution.png",
       plot = daa_p_adj_summary_plot, width = 10, height = 6)

ggsave("../visualizations/figures/functional_analysis/log2fc_distribution.png",
       plot = daa_log2fc_summary_plot, width = 10, height = 6)
```
