---
title: "functional_potential"
output: html_document
creator: "Alex Elbaz"
---

# Predicted Functional Potential Analyses (PICRUSt2)

## 1. Load Libraries

```{r}
library(tidyverse)
library(phyloseq)
library(ggpicrust2)
library(devtools)
install_github("biobakery/maaslin2")

# A log2 fold change calculation function alongside a modified ggpicrust2 errorbar function, both from the MICB 305 2025W2 GitHub, are used within this script (imports below).
# This code is attributed to the creators of the MICB 305 2025W2 GitHub and is cited within our research paper.
source('reference_functions/ggpicrust2_calculate_log2fc.R')
source('reference_functions/ggpicrust2_errorbar_function_fixed.R')
```

## 2. Load Data + Initial Metadata Wrangling

Functional annotations were generated via command-line PICRUSt2 and then exported into this project under the `../processed_data/` directory.

```{r}
# Load in metadata & functional pathway data
metadata = data.frame(readRDS('../processed_data/phyloseq_taxonomy.rds')@sam_data) %>%
  rownames_to_column()

metacyc_path_abun = read.delim('../processed_data/path_abun_unstrat.tsv', row.names = 1)


# Rename and order infection status groups so the downstream figures are more interpretable
metadata$influenza_status = metadata$influenza_status %>%
  replace_values(from = "Pos", to = "infected") %>%
  replace_values(from = "Neg", to = "uninfected") %>%
  factor(levels = c("infected", "uninfected"))

# Generate metadata data frame with uninfected samples filtered out for HA Group & NA Group analyses. Also, rename grouping variable so the downstream figures are more interpretable
metadata_only_infected = filter(metadata, ha_group != "neg" & na_group != "neg")
metadata_only_infected$ha_group = ifelse(metadata_only_infected$ha_group == "1",
                                         "HA Group 1",
                                         "HA Group 2")
metadata_only_infected$na_group = ifelse(metadata_only_infected$na_group == "1",
                                         "NA Group 1",
                                         "NA Group 2")

# Generate metacyc pathway abundance data frame with uninfected samples filtered out for HA Group & NA Group analyses
metacyc_path_abun_only_infected = metacyc_path_abun[, colnames(metacyc_path_abun) %in% metadata_only_infected$rowname]
```

## 3. Set Seed

```{r}
set.seed(1234)
```

## 4. Conduct Functional Potential Analysis (via ggpicrust2)

This analysis examines differences in predicted functional potential between uninfected and influenza-infected mallards, HA Group 1 & HA Group 2 influenza-infected mallards, and NA Group 1 & NA Group 2 influenza-infected mallards.

### 4.1 Infected vs. Uninfected

```{r}
# Run pathway differential abundance analysis (Maaslin2 method)
daa_infection = pathway_daa(abundance = metacyc_path_abun,
                     metadata = metadata,
                     group = "influenza_status",
                     daa_method = "Maaslin2",
                     select = NULL,
                     reference = NULL) %>%
  select(-log2_fold_change)

# Get correct log2 fold change values; currently, pathway_daa using the Maaslin2 method uses incorrect log2 fold changes.
log2fc_infection = fix_lfc(abundance = metacyc_path_abun,
                           metadata = metadata,
                           group = "influenza_status",
                           reference = "uninfected") %>%
  select(feature, log2_fold_change)

# Replace incorrect log2 fold change values with the correct ones
daa_infection = left_join(daa_infection, log2fc_infection, by="feature") %>%
  relocate(log2_fold_change, .before = adj_method)


# Add pathway annotations
daa_infection = pathway_annotation(pathway = "MetaCyc", daa_results_df = daa_infection)
```

### 4.2 HA Group 1 vs. HA Group 2

```{r}
# Run pathway differential abundance analysis (Maaslin2 method)
daa_ha = pathway_daa(abundance = metacyc_path_abun_only_infected,
                     metadata = metadata_only_infected,
                     group = "ha_group",
                     daa_method = "Maaslin2",
                     select = NULL,
                     reference = NULL) %>%
  select(-log2_fold_change)

# Get correct log2 fold change values; currently, pathway_daa using the Maaslin2 method uses incorrect log2 fold changes.
log2fc_ha = fix_lfc(abundance = metacyc_path_abun_only_infected,
                         metadata = metadata_only_infected,
                         group = "ha_group",
                         reference = "HA Group 2") %>%
  select(feature, log2_fold_change)

# Replace incorrect log2 fold change values with the correct ones
daa_ha = left_join(daa_ha, log2fc_ha, by="feature") %>%
  relocate(log2_fold_change, .before = adj_method)

# Add pathway annotations
daa_ha = pathway_annotation(pathway = "MetaCyc", daa_results_df = daa_ha)
```

### 4.3 NA Group 1 vs. NA Group 2

```{r}
# Run pathway differential abundance analysis (Maaslin2 method)
daa_na = pathway_daa(abundance = metacyc_path_abun_only_infected,
                     metadata = metadata_only_infected,
                     group = "na_group",
                     daa_method = "Maaslin2",
                     select = NULL,
                     reference = NULL) %>%
  select(-log2_fold_change)

# Get correct log2 fold change values; currently, pathway_daa using the Maaslin2 method uses incorrect log2 fold changes.
log2fc_na = fix_lfc(abundance = metacyc_path_abun_only_infected,
                         metadata = metadata_only_infected,
                         group = "na_group",
                         reference = "NA Group 2") %>%
  select(feature, log2_fold_change)

# Replace incorrect log2 fold change values with the correct ones
daa_na = left_join(daa_na, log2fc_na, by="feature") %>%
  relocate(log2_fold_change, .before = adj_method)

# Add pathway annotations
daa_na = pathway_annotation(pathway = "MetaCyc", daa_results_df = daa_na)
```

## 5. Generate Figures (Bar plots with most significant pathways)

### 5.1 Infected vs. Uninfected

An adjusted p-value cutoff of \< 0.05 and absolute log2 fold change cutoff of \>= 2 was applied in order to highlight the most significant results while maintaining figure readability.

```{r}
# Create column containing both pathway names & annotations
daa_infection = daa_infection %>%
  mutate(label = paste(feature, ifelse(is.na(description),
                                       "No description available",
                                       description),
                       sep = "; "))

# Filter for significant results
daa_infection_filt = filter(daa_infection, abs(log2_fold_change) >= 2 & p_adjust < 0.05)

# Generate pathway errorbar plot
peb_infection = pathway_errorbar_fixed(abundance = metacyc_path_abun,
                                       daa_results_df = daa_infection_filt,
                                       Group = metadata$influenza_status,
                                       order_by_log = TRUE,
                                       ko_to_kegg = FALSE,
                                       p_value_bar = TRUE,
                                       x_lab = "label")
```

### 5.2 HA Group 1 vs. HA Group 2

There were no significant comparisons between HA groups. However, in order to produce a readable figure and highlight the comparisons with the lowest p-values and largest difference between groups, an adjusted p-value cutoff of \< 0.95 alongside an absolute log2 fold change cutoff of \>= 2.5 was applied.

```{r}
# Create column containing both pathway names & annotations
daa_ha = daa_ha %>%
  mutate(label = paste(feature, ifelse(is.na(description),
                                       "No description available",
                                       description),
                       sep = "; "))

# Filter for significant results
daa_ha_filt = filter(daa_ha, abs(log2_fold_change) >= 2.5 & p_adjust < 0.95)

# Generate pathway errorbar plot
peb_ha = pathway_errorbar_fixed(abundance = metacyc_path_abun_only_infected,
                             daa_results_df = daa_ha_filt,
                             Group = metadata_only_infected$ha_group,
                             order_by_log = TRUE,
                             ko_to_kegg = FALSE,
                             p_value_bar = TRUE,
                             p_values_threshold = 1,
                             x_lab = "label")
```

### 5.3 NA Group 1 vs. NA Group 2

There were no significant comparisons between NA groups. However, in order to produce a readable figure and highlight the comparisons with the lowest p-values and largest difference between groups, an adjusted p-value cutoff of \< 0.65 alongside an absolute log2 fold change cutoff of \>= 2.5 was applied.

```{r}
# Create column containing both pathway names & annotations
daa_na = daa_na %>%
  mutate(label = paste(feature, ifelse(is.na(description),
                                       "No description available",
                                       description),
                       sep = "; "))

# Filter for significant results
daa_na_filt = filter(daa_na, abs(log2_fold_change) >= 2.5 & p_adjust < 0.65)

# Generate pathway errorbar plot
peb_na = pathway_errorbar_fixed(abundance = metacyc_path_abun_only_infected,
                             daa_results_df = daa_na_filt,
                             Group = metadata_only_infected$na_group,
                             order_by_log = TRUE,
                             ko_to_kegg = FALSE,
                             p_value_bar = TRUE,
                             p_values_threshold = 1,
                             x_lab = "label")
```

## 6. Generate Significance & Effect Size Summary Plots

### 6.1 Adjusted P-value Distributions by Comparison

```{r}
# Aggregate adjusted p-value data into a single data frame
daa_p_adj_summary = data.frame("Uninfected_vs._Infected" = daa_infection$p_adjust,
                               "HA_Group_1_vs._HA_Group_2" = daa_ha$p_adjust,
                               "NA_Group_1_vs._NA_Group_2" = daa_na$p_adjust)

# Pivot aggregated data, repair group names, remove NAs, & order groups as desired for plotting
daa_p_adj_summary = pivot_longer(daa_p_adj_summary,
                                 cols = everything(),
                                 names_to = "group",
                                 values_to = "p_adj") %>%
  na.omit()

daa_p_adj_summary$group = sapply(daa_p_adj_summary$group,
                                 str_replace_all,
                                 pattern = "_",
                                 replacement = " ") %>%
  factor(levels = c("Uninfected vs. Infected",
                    "HA Group 1 vs. HA Group 2",
                    "NA Group 1 vs. NA Group 2"))


# Generate adjusted p-value distribution summary plot
daa_p_adj_summary_plot = ggplot(daa_p_adj_summary, aes(x = p_adj)) +
  geom_histogram(binwidth = 0.02, colour = "black", fill = "lightgrey", linewidth = 0.2) +
  facet_wrap(~group, ncol = 1, scales="free_y") +
  labs(x = "Adjusted P value", y = "Count") +
  theme_bw() +
  theme(text = element_text(size = 14, hjust = 0.5))
# Mark the means in red on the histograms?
```

### 6.2 Log2 Fold Change Distributions by Comparison

```{r}
# Aggregate adjusted p-value data into a single data frame
daa_log2fc_summary = data.frame("Uninfected_vs._Infected" = daa_infection$log2_fold_change,
                                "HA_Group_1_vs._HA_Group_2" = daa_ha$log2_fold_change,
                                "NA_Group_1_vs._NA_Group_2" = daa_na$log2_fold_change)

# Pivot aggregated data, repair group names, remove NAs, & order groups as desired for plotting
daa_log2fc_summary = pivot_longer(daa_log2fc_summary,
                                  cols = everything(),
                                  names_to = "group",
                                  values_to = "log2fc") %>%
  na.omit()

daa_log2fc_summary$group = sapply(daa_log2fc_summary$group,
                                  str_replace_all,
                                  pattern = "_",
                                  replacement = " ") %>%
  factor(levels = c("Uninfected vs. Infected",
                    "HA Group 1 vs. HA Group 2",
                    "NA Group 1 vs. NA Group 2"))


# Generate log2 fold change distribution summary plot
daa_log2fc_summary_plot = ggplot(daa_log2fc_summary, aes(x = log2fc)) +
  geom_histogram(binwidth = 0.25, colour = "black", fill = "lightgrey", linewidth = 0.2) +
  facet_wrap(~group, ncol = 1, scales="free_y") +
  labs(x = "Log2 Fold Change", y = "Count") +
  theme_bw() +
  theme(text = element_text(size = 14, hjust = 0.5))
# Mark the means in red on the histograms?
```

## 7. Save Figures

```{r}
# Functional potential figures
ggsave("../visualizations/figures/functional_analysis/functional_potential_control_infected.png", plot = peb_infection, width = 18, height = 8)

ggsave("../visualizations/figures/functional_analysis/functional_potential_ha_groups.png",
       plot = peb_ha, width = 18, height = 8)

ggsave("../visualizations/figures/functional_analysis/functional_potential_na_groups.png",
       plot = peb_na, width = 18, height = 8)


# Significance & effect size summary figures
ggsave("../visualizations/figures/functional_analysis/p_adj_distribution_summary.png",
       plot = daa_p_adj_summary_plot, width = 10, height = 6)

ggsave("../visualizations/figures/functional_analysis/log2fc_distribution_summary.png",
       plot = daa_log2fc_summary_plot, width = 10, height = 6)
```
