---
title: "functional_potential"
output: html_document
creator: "Alex Elbaz"
---

# Predicted Functional Potential Analyses (PICRUSt2)

## 1. Load Libraries

```{r}
library(tidyverse)
library(phyloseq)
library(ggpicrust2)
library(devtools)
install_github("zhouhj1994/LinDA")

# A modified ggpicrust2 errorbar function from the MICB 305 2025W2 GitHub is used within this script (import below).
# This code is attributed to the creators of the MICB 305 2025W2 GitHub and is cited within our research paper.
source('reference_functions/ggpicrust2_errorbar_function_fixed.R')
```

## 2. Load Data

Functional annotations were generated via command-line PICRUSt2 and then exported into this project under the `../processed_data/` directory.

```{r}
metadata = read.delim('../processed_data/metadata_processed.tsv')
metacyc_path_abun = read.delim('../processed_data/path_abun_unstrat.tsv', row.names = 1)
```

## 3. Set Seed

```{r}
set.seed(1234)
```

## 4. Conduct Functional Potential Analysis (via ggpicrust2)

This analysis examines differences in predicted functional potential between infected and uninfected mallards.

```{r}
# Rename and order groups appropriately so the downstream figure is more interpretable
metadata$influenza_status = replace_values(metadata$influenza_status, from = "Pos", to = "infected") %>%
  replace_values(from = "Neg", to = "uninfected") %>%
  factor(levels = c("infected", "uninfected"))

# Run pathway differential abundance analysis (LinDA method)
daa_df = pathway_daa(abundance = metacyc_path_abun,
                     metadata = metadata,
                     group = "influenza_status",
                     daa_method = "LinDA",
                     select = NULL,
                     reference = NULL)

# Add pathway annotations
daa_annotated_df = pathway_annotation(pathway = "MetaCyc", daa_results_df = daa_df)
```

## 5. Generate Figure (Bar plot showing significant pathways)

Of note, an adjusted p-value cutoff of \< 0.001 and absolute log2 fold change cutoff of \>= 2.5 was applied in order to highlight the most significant results while maintaining figure readability.

```{r}
# Create column containing both pathway names & annotations
daa_annotated_df = daa_annotated_df %>%
  mutate(label = paste(feature, ifelse(is.na(description), "No description available", description), sep = "; "))

# Filter for significant results
daa_annotated_df_filt = filter(daa_annotated_df, abs(log2FoldChange) >= 2.5 & p_adjust < 0.001)

# Generate pathway errorbar plot
peb = pathway_errorbar_fixed(abundance = metacyc_path_abun,
                             daa_results_df = daa_annotated_df_filt,
                             Group = metadata$influenza_status,
                             order_by_log = TRUE,
                             ko_to_kegg = FALSE,
                             p_value_bar = TRUE,
                             x_lab = "label")
```

## 6. Save Figure

```{r}
ggsave("../visualizations/figures/functional_potential.png", plot = peb, width = 18, height = 10)
```

## 7. Additional Useful Information

```{r}
num_pathways = nrow(daa_annotated_df)
num_signif = nrow(filter(daa_annotated_df, p_adjust < 0.05))
prop_signif = num_signif/num_pathways

print(paste("Pathways analyzed:", num_pathways))
print(paste("Significant pahtways (p_adjust < 0.05):", num_signif))
print(paste("Proportion significant:", round(prop_signif, 2)))
```
