---
title: "functional_potential"
output: html_document
creator: "Alex Elbaz"
---

# Predicted Functional Potential Analyses (PICRUSt2)

## 1. Load Libraries

```{r}
library(tidyverse)
library(phyloseq)
library(ggpicrust2)
library(devtools)
install_github("biobakery/maaslin2")

# A log2 fold change calculation function alongside a modified ggpicrust2 errorbar function, both from the MICB 305 2025W2 GitHub, are used within this script (imports below).
# This code is attributed to the creators of the MICB 305 2025W2 GitHub and is cited within our research paper.
source('reference_functions/ggpicrust2_calculate_log2fc.R')
source('reference_functions/ggpicrust2_errorbar_function_fixed.R')

# Some internal functions appear inaccessible in ggpicrust2, so I am importing the internal functions separately for now
source('reference_functions/data_utils.R')
```

## 2. Load Data

Functional annotations were generated via command-line PICRUSt2 and then exported into this project under the `../processed_data/` directory.

```{r}
metadata = data.frame(readRDS('../processed_data/phyloseq_taxonomy.rds')@sam_data) %>%
  rownames_to_column()

metacyc_path_abun = read.delim('../processed_data/path_abun_unstrat.tsv', row.names = 1)
```

## 3. Set Seed

```{r}
set.seed(1234)
```

## 4. Conduct Functional Potential Analysis (via ggpicrust2)

This analysis examines differences in predicted functional potential between uninfected and influenza-infected mallards as well as between HA group 1 and HA group 2 influenza-infected mallards.

### 4.1 Infected vs. Uninfected

```{r}
# Rename and order groups appropriately so the downstream figure is more interpretable
metadata$influenza_status = replace_values(metadata$influenza_status, from = "Pos", to = "infected") %>%
  replace_values(from = "Neg", to = "uninfected") %>%
  factor(levels = c("infected", "uninfected"))


# Run pathway differential abundance analysis (Maaslin2 method)
daa_ctrl_inf = pathway_daa(abundance = metacyc_path_abun,
                     metadata = metadata,
                     group = "influenza_status",
                     daa_method = "Maaslin2",
                     select = NULL,
                     reference = NULL) %>%
  select(-log2_fold_change)

# Get correct log2 fold change values; currently, pathway_daa using the Maaslin2 method uses incorrect log2 fold changes.
log2_fc_ctrl_inf = fix_lfc(abundance = metacyc_path_abun,
                           metadata = metadata,
                           group = "influenza_status",
                           reference = "uninfected") %>%
  select(feature, log2_fold_change)

# Replace incorrect log2 fold change values with the correct ones
daa_ctrl_inf = left_join(daa_ctrl_inf, log2_fc_ctrl_inf, by="feature") %>%
  relocate(log2_fold_change, .before = adj_method)


# Add pathway annotations
daa_ctrl_inf_annotated = pathway_annotation(pathway = "MetaCyc", daa_results_df = daa_ctrl_inf)
```

### 4.2 HA Group 1 vs. HA Group 2

```{r}
# Filter out uninfected samples and rename groups so downstream figure is more interpretable
metadata_filt = filter(metadata, ha_group != "neg")
metadata_filt$ha_group = paste("ha_group", metadata_filt$ha_group, sep="_")

metacyc_path_abun_filt = metacyc_path_abun[, colnames(metacyc_path_abun) %in% metadata_filt$rowname]


# Run pathway differential abundance analysis (Maaslin2 method)
daa_ha_grp = pathway_daa(abundance = metacyc_path_abun_filt,
                     metadata = metadata_filt,
                     group = "ha_group",
                     daa_method = "Maaslin2",
                     select = NULL,
                     reference = NULL) %>%
  select(-log2_fold_change)

# Get correct log2 fold change values; currently, pathway_daa using the Maaslin2 method uses incorrect log2 fold changes.
log2_fc_ha_grp = fix_lfc(abundance = metacyc_path_abun_filt,
                         metadata = metadata_filt,
                         group = "ha_group",
                         reference = "ha_group_2") %>%
  select(feature, log2_fold_change)

# Replace incorrect log2 fold change values with the correct ones
daa_ha_grp = left_join(daa_ha_grp, log2_fc_ha_grp, by="feature") %>%
  relocate(log2_fold_change, .before = adj_method)

# Add pathway annotations
daa_ha_grp_annotated = pathway_annotation(pathway = "MetaCyc", daa_results_df = daa_ha_grp)
```

## 5. Generate Figure (Bar plot with most significant pathways)

### 5.1 Infected vs. Uninfected

An adjusted p-value cutoff of \< 0.05 and absolute log2 fold change cutoff of \>= 2 was applied in order to highlight the most significant results while maintaining figure readability.

```{r}
# Create column containing both pathway names & annotations
daa_ctrl_inf_annotated = daa_ctrl_inf_annotated %>%
  mutate(label = paste(feature, ifelse(is.na(description), "No description available", description), sep = "; "))

# Filter for significant results
daa_ctrl_inf_annotated_filt = filter(daa_ctrl_inf_annotated,
                                     abs(log2_fold_change) >= 2 & p_adjust < 0.05)

# Generate pathway errorbar plot
peb_ctrl_inf = pathway_errorbar_fixed(abundance = metacyc_path_abun,
                             daa_results_df = daa_ctrl_inf_annotated_filt,
                             Group = metadata$influenza_status,
                             order_by_log = TRUE,
                             ko_to_kegg = FALSE,
                             p_value_bar = TRUE,
                             x_lab = "label")
```

### 5.2 HA Group 1 vs. HA Group 2

There were no significant comparisons between HA groups. However, in order to produce a readable figure and highlight the comparisons with the lowest p-values and largest difference between groups, an adjusted p-value cutoff of \< 0.9 alongside an absolute log2 fold change cutoff of \>= 2 was applied.

```{r}
# Create column containing both pathway names & annotations
daa_ha_grp_annotated = daa_ha_grp_annotated %>%
  mutate(label = paste(feature, ifelse(is.na(description), "No description available", description), sep = "; "))

# Filter for significant results
daa_ha_grp_annotated_filt = filter(daa_ha_grp_annotated,
                                   abs(log2_fold_change) >= 2 & p_adjust < 0.9)

# Generate pathway errorbar plot
peb_ha_grp = pathway_errorbar_fixed(abundance = metacyc_path_abun_filt,
                             daa_results_df = daa_ha_grp_annotated_filt,
                             Group = metadata_filt$ha_group,
                             order_by_log = TRUE,
                             ko_to_kegg = FALSE,
                             p_value_bar = TRUE,
                             p_values_threshold = 1,
                             x_lab = "label")
```

## 6. Save Figure

```{r}
ggsave("../visualizations/figures/functional_potential_control_infected.png",
       plot = peb_ctrl_inf, width = 18, height = 8)

ggsave("../visualizations/figures/functional_potential_ha_group.png",
       plot = peb_ha_grp, width = 18, height = 10)
```

## 7. Additional Useful Information

### 7.1 Infected vs. Uninfected

```{r}
num_pathways = nrow(daa_ctrl_inf_annotated)
num_signif = nrow(filter(daa_ctrl_inf_annotated, p_adjust < 0.05))
prop_signif = num_signif/num_pathways

print(paste("Pathways analyzed:", num_pathways))
print(paste("Significant pahtways (p_adjust < 0.05):", num_signif))
print(paste("Proportion significant:", round(prop_signif, 2)))
```

### 7.2 HA Group 1 vs. HA Group 2

```{r}
num_pathways = nrow(daa_ha_grp_annotated)
num_signif = nrow(filter(daa_ha_grp_annotated, p_adjust < 0.05))
prop_signif = num_signif/num_pathways

print(paste("Pathways analyzed:", num_pathways))
print(paste("Significant pahtways (p_adjust < 0.05):", num_signif))
print(paste("Proportion significant:", round(prop_signif, 2)))
```
