---
title: "generate_phyloseq"
output: html_document
---

# Generate Phyloseq Object for use in Analyses

## 1. Load Libraries

```{r}
library(tidyverse)
library(phyloseq)
```

## 2. Load Data

```{r}
metadata = read.delim('../processed_data/metadata_processed.tsv', row.names = 1)
taxonomy = read.delim('../processed_data/taxonomy.tsv', row.names = 1)
counts = read.delim('../processed_data/feature-table.txt', skip=1, row.names=1)
tree = read_tree("../processed_data/tree.nwk")
```

## 3. Wrangle Metadata (i.e., create the groupings used for analysis)

For the purposes of our analyses, we deemed it appropriate analyze HA & NA groups separately. Additionally, given the small sample size of our data, we deemed it most appropriate to aggregate HA and NA groups as follows:

-   Both HA & NA: Aggregate sub-subtypes to subtypes (i.e., H5.2 --\> H5, N3.2 --\> N3, etc.).

-   HA: Bucket subtypes into two groups based on protein phylogeny established in prior literature.

```{r}
# HA Group 1 includes H1, H2, H5, H6, H8, H9, H11, H12, H13, & H16
ha_grp_1 = c("1", "2", "5", "6", "8", "9", "11", "12", "13", "16")

# HA Group 2 includes H3, H4, H7, H10, H14, & H15
ha_grp_2 = c("3", "4", "7", "10", "14", "15")

metadata = metadata %>%
  mutate(ha_aggr = str_extract(ha_subtype, "^[^.]+")) %>%
  mutate(na_aggr = str_extract(na_subtype, "^[^.]+")) %>%
  mutate(ha_group = ifelse(ha_aggr %in% ha_grp_1, "1", ifelse(ha_aggr %in% ha_grp_2, "2", "neg")))
```

## 4. Generate Phyloseq Object

```{r}
# Format Taxonomy
taxa_formatted = taxonomy %>%
  separate(col = Taxon,
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
           fill="right",
           sep = ";") %>%
  select(-Confidence) %>%
  as.matrix()

# Format counts
counts_formatted = counts %>% as.matrix()

# Generate Phyloseq object
ps = phyloseq(sample_data(metadata),
              otu_table(counts_formatted, taxa_are_rows = TRUE),
              tax_table(taxa_formatted),
              tree)
```

## 5. Save Phyloseq Object

```{r}
saveRDS(ps, "../processed_data/phyloseq_taxonomy.rds")
```
