---
title: "indicator_taxa"
output: html_document
creator: "Alex Elbaz"
---

# Indicator Taxa Analysis

## 1. Load Libraries

```{r}
library(tidyverse)
library(phyloseq)
library(vegan)
library(ggpubr)
library(microbiome)
library(indicspecies)
library(gridExtra)
```

## 2. Load Data

For the purposes of this analysis, we deemed it appropriate analyze HA & NA groups separately. Additionally, to increase statistical power (given our small sample size), we decided to aggregate sub-subtypes to subtypes (i.e., H5.2 --\> H5, N3.2 --\> N3, etc.).

```{r}
# Read in phyloseq object
ps = readRDS('../processed_data/phyloseq_taxonomy.rds')

# Mutate metadata as described above
new_metadata = data.frame(ps@sam_data) %>%
  mutate(ha_subtype = str_extract(ha_subtype, "^[^.]+")) %>%
  mutate(na_subtype = str_extract(na_subtype, "^[^.]+"))

# Insert mutated metadata into the phyloseq object
ps@sam_data = sample_data(new_metadata)
```

## 3. Set Seed & Prepare Data for Analyses

We deemed it most appropriate to agglomerate to the genus level for this analysis given the level of annotation available in the taxonomy data.

```{r}
# Set seed for repeatable analysis
set.seed(1234)

# Agglomerate taxa to genus level, convert to relative abundance, & apply abundance filter
ps_filt = tax_glom(ps, "Genus") %>%
  microbiome::transform("compositional") %>%
  filter_taxa(function(x) mean(x) >= 0.0001, TRUE)
```

## 4. Conduct Indicator Taxa Analysis (via indicpecies package)

```{r}
# Conduct indicator taxa analysis for HA groups & filter results for stat & p-value cutoffs
ita_ha = multipatt(t(otu_table(ps_filt)), 
                       cluster = sample_data(ps_filt)$ha_subtype,
                       control = how(nperm = 999))$sign %>%
  as.data.frame(.) %>%
  filter(stat > 0.7 & p.value < 0.05)


# Conduct indicator taxa analysis for NA groups & filter results for stat & p-value cutoffs
ita_na = multipatt(t(otu_table(ps_filt)), 
                       cluster = sample_data(ps_filt)$na_subtype,
                       control = how(nperm = 999))$sign %>%
  as.data.frame(.) %>%
  filter(stat > 0.7 & p.value < 0.05)
```

## 5. Generate Indicator Taxa Summary Tables for HA & NA Analyses

The tables are organized such that each row corresponds to an indicator genus. The values of the columns H2-H5 (or N1-N9 excluding N4) and neg represent whether the given genus is an indicator for that group (1) or not (0). Additionally, the tables are filtered to only include significant results (p \< 0.05 & stat \> 0.7).

### 5.1 HA Table

```{r}
# Get indicator genera names from indicator genera ASVs
indicator_genera_ha = data.frame(ps_filt@tax_table) %>%
  rownames_to_column() %>%
  filter(rowname %in% rownames(ita_ha)) %>%
  select(rowname, Genus)
indicator_genera_ha$Genus = indicator_genera_ha$Genus %>% str_replace("g__", "")

# Generate summary table
ita_summary_ha = ita_ha %>%
  rownames_to_column() %>%
  left_join(indicator_genera_ha)

# Format summary table for ease of interpretation
ita_summary_ha_formatted = ita_summary_ha %>%
  rename(H2 = "s.2", H3 = "s.3", H4 = "s.4", H5 = "s.5", neg = "s.neg") %>%
  select(-index, -rowname) %>%
  relocate(stat, .after = last_col()) %>%
  relocate(p.value, .before = stat) %>%
  relocate(Genus, .before = H2) %>%
  mutate(stat = round(stat, 3)) %>%
  rename(indicator_value = stat, p_value = p.value)

# Display formatted summary table
ita_summary_ha_formatted
```

### 5.2 NA Table

```{r}
# Get indicator genera names from indicator genera ASVs
indicator_genera_na = data.frame(ps_filt@tax_table) %>%
  rownames_to_column() %>%
  filter(rowname %in% rownames(ita_na)) %>%
  select(rowname, Genus)
indicator_genera_na$Genus = indicator_genera_na$Genus %>% str_replace("g__", "")

# Generate summary table
ita_summary_na = ita_na %>%
  rownames_to_column() %>%
  left_join(indicator_genera_na)

# Format summary table for ease of interpretation
ita_summary_na_formatted = ita_summary_na %>%
  rename(N1 = "s.1", N2 = "s.2", N3 = "s.3", N5 = "s.5",
         N6 = "s.6", N7 = "s.7", N8 = "s.8", N9 = "s.9", neg = "s.neg") %>%
  select(-index, -rowname) %>%
  relocate(stat, .after = last_col()) %>%
  relocate(p.value, .before = stat) %>%
  relocate(Genus, .before = N1) %>%
  mutate(stat = round(stat, 3)) %>%
  rename(indicator_value = stat, p_value = p.value)

# Display formatted summary table
ita_summary_na_formatted
```

### 5.4 Save HA & NA Tables

```{r}
pdf("../visualizations/tables/ita_ha_summary.pdf", paper = "USr")
grid.table(ita_summary_ha_formatted, theme = ttheme_default(base_size = 8), rows = NULL)
dev.off()

pdf("../visualizations/tables/ita_na_summary.pdf", paper = "USr")
grid.table(ita_summary_na_formatted, theme = ttheme_default(base_size = 8), rows = NULL)
dev.off()
```
