---
title: "indicator_taxa"
output: html_document
author: "Alex Elbaz"
---

# Indicator Taxa Analysis

## 1. Load Libraries

```{r}
library(tidyverse)
library(phyloseq)
library(vegan)
library(ggpubr)
library(microbiome)
library(indicspecies)
library(gridExtra)
library(gtable)
library(grid)
```

## 2. Define Functions

```{r}
# Agglomerate taxa to genus level, convert to relative abundance, & apply relative abundance filter
# ps: phyloseq object
# abun_thresh: abundance filter threshold value
ps_ita_transform = function(ps, abun_thresh = 0.0001) {
  return(tax_glom(ps, "Genus") %>%
           microbiome::transform("compositional") %>%
           filter_taxa(function(x) mean(x) >= abun_thresh, TRUE))
}


# Conduct indicator taxa analysis
# ps: phyloseq object
# group: metadata column defining the groups of interest
ita = function(ps, group) {
  return(multipatt(t(otu_table(ps)), 
                   cluster = sample_data(ps)[[group]],
                   control = how(nperm = 999))$sign %>%
           as.data.frame(.))
}


# Generate formatted indicator taxa summary table
# ps: phyloseq object
# ita_results: data frame containing ita results
# grp1_name: name for column corresponding to group 1
# grp2_name: name for column corresponding to group 2
# uninfected_name: name for column corresponding to uninfected samples (or NULL if no uninfected samples included)
format_summary_table = function(ps, ita_results, grp1_name = "s.1", grp2_name = "s.2", uninfected_name = NULL) {
  # Get indicator genera names from indicator genera ASVs
  indic_genera = data.frame(ps@tax_table) %>%
    rownames_to_column() %>%
    filter(rowname %in% rownames(ita_results)) %>%
    select(rowname, Genus)
  
  indic_genera$Genus = indic_genera$Genus %>% str_replace("g__", "")
  
  # Generate summary table
  ita_summary = ita_results %>%
    rownames_to_column() %>%
    left_join(indic_genera)
  
  # Format summary table for ease of interpretation
  cols = colnames(ita_summary)
  
  if (!is.null(grp1_name)) {
    cols = replace(cols, cols=="s.1", grp1_name)
  }
  
  if (!is.null(grp2_name)) {
    cols = replace(cols, cols=="s.2", grp2_name)
  }
  
  if (!is.null(uninfected_name)) {
    cols = replace(cols, cols=="s.neg", uninfected_name)
  }
  
  ita_summary_formatted = ita_summary %>%
    `colnames<-`(cols) %>%
    select(-index, -rowname) %>%
    relocate(stat, .after = last_col()) %>%
    relocate(p.value, .before = stat) %>%
    relocate(Genus, .before = grp1_name) %>%
    mutate(stat = round(stat, 3)) %>%
    mutate(p.value = round(p.value, 3)) %>%
    rename("Indicator Value" = stat, "P Value (Adjusted)" = p.value)
  
  # Re-label indicator 0/1 markers with ✘/✔
  for(col in c(grp1_name, grp2_name, uninfected_name)) {
    if (!is.null(col)) {
      ita_summary_formatted[[col]] = replace(ita_summary_formatted[[col]],
                                             which(1 == ita_summary_formatted[[col]]),
                                             "✔")
      ita_summary_formatted[[col]] = replace(ita_summary_formatted[[col]],
                                             which(0 == ita_summary_formatted[[col]]),
                                             "✘")
    }
  }
  
  return(ita_summary_formatted)
}


# Save indicator taxa analysis table formatted appropriately.
# tbl: indicator taxa results table
# name: name for the output table file
save_ita_tbl = function(tbl, name) {
  # Order table rows consistently prior to exporting the figure
  tbl = tbl %>% arrange(Genus, desc("Indicator Value"))

  # Export figure
  cairo_pdf(paste("../visualizations/tables/indicator_taxa/", name, ".pdf", sep = ""))
  tbl_grob = tableGrob(tbl, theme = ttheme_minimal(base_size = 8), rows = NULL) %>%
    gtable_add_grob(grobs = segmentsGrob(x0 = unit(0, "npc"),
                                         y0 = unit(0, "npc"),
                                         x1 = unit(1, "npc"),
                                         y1 = unit(0, "npc")),
                    t = 1, l = 1, b = 1, r = ncol(tbl))
  grid.newpage()
  grid.draw(tbl_grob)
  dev.off()
}
```

## 3. Load Data

```{r}
# Read in phyloseq object; keep all samples
ps_all = readRDS('../processed_data/phyloseq_taxonomy.rds')

# Read in phyloseq object; remove uninfected samples
ps_infected = readRDS('../processed_data/phyloseq_taxonomy.rds')
ps_infected@sam_data = data.frame(ps_infected@sam_data) %>%
  filter(ha_group != "neg" & na_group != "neg") %>%
  sample_data(.)
```

## 4. Set Seed & Prepare Data for Analyses

We deemed it most appropriate to agglomerate to the genus level for this analysis given the level of annotation available in the taxonomy data.

```{r}
# Set seed for repeatable analysis
set.seed(1234)

# Agglomerate taxa to genus level, convert to relative abundance, & apply relative abundance filter of 0.0001
ps_filt_all = ps_ita_transform(ps_all, 0.0001)
ps_filt_infected = ps_ita_transform(ps_infected, 0.0001)
```

## 5. Conduct Indicator Taxa Analysis (via indicpecies package)

```{r}
# HA Groups including uninfected samples
ita_ha_all = ita(ps_filt_all, "ha_group")

# NA Groups including uninfected samples
ita_na_all = ita(ps_filt_all, "na_group")

# HA Groups without uninfected samples
ita_ha_infected = ita(ps_filt_infected, "ha_group") 
  
# NA Groups without uninfected samples
ita_na_infected = ita(ps_filt_infected, "na_group")
```

## 6. Adjust P Values for Multiple Comparisons

There is no indication that the multipatt() function adjusts p-values to account for multiple comparisons. Additionally, two statistical analyses were conducted for each set of unique data (i.e., two using HA group data and two using NA group data). Hence, a p-value adjustment (Benjamini-Hochberg procedure) will be applied to adjust the p-values for both sets of comparisons.

### 5.1 HA Analyses p-adjustment

```{r}
num_ha_comparisons = length(ita_ha_all$p.value) + length(ita_ha_infected$p.value)
ita_ha_all$p.value = p.adjust(ita_ha_all$p.value, method = "BH", n = num_ha_comparisons)
ita_ha_infected$p.value = p.adjust(ita_ha_infected$p.value, method = "BH", n = num_ha_comparisons)
```

### 5.2 NA Analyses p-adjustment

```{r}
num_na_comparisons = length(ita_na_all$p.value) + length(ita_na_infected$p.value)
ita_na_all$p.value = p.adjust(ita_na_all$p.value, method = "BH", n = num_na_comparisons)
ita_na_infected$p.value = p.adjust(ita_na_infected$p.value, method = "BH", n = num_na_comparisons)
```

## 7. Remove NAs & Filter for Significant Results

Notably, the threshold used to establish significance was adjusted p \< 0.05 and indicator value \> 0.7.

```{r}
ita_ha_all = na.omit(ita_ha_all)
ita_ha_infected = na.omit(ita_ha_infected)
ita_na_all = na.omit(ita_na_all)
ita_na_infected = na.omit(ita_na_infected)

ita_ha_all_significant = ita_ha_all %>% filter(p.value < 0.05 & stat > 0.7)
ita_ha_infected_significant = ita_ha_infected %>% filter(p.value < 0.05 & stat > 0.7)
ita_na_all_significant = ita_na_all %>% filter(p.value < 0.05 & stat > 0.7)
ita_na_infected_significant = ita_na_infected %>% filter(p.value < 0.05 & stat > 0.7)
```

## 8. Generate Indicator Taxa Summary Tables

The tables are organized such that each row corresponds to an indicator genus. The values of the columns "HA Group 1", "HA Group 2", "NA Group 1", "NA Group 2", and "Uninfected" represent whether the given genus is an indicator for that group (✔) or not (✘).

Importantly, for HA & NA analyses including uninfected samples, only significant taxa are included to produce a readable figure (stat \> 0.7 & adjusted p \< 0.05).

Conversely, there were **NO** significant indicator taxa found for HA & NA analyses without uninfected samples, so the top 10 most significant taxa are shown in the respective tables; uncultured taxa are excluded.

```{r}
# HA Groups including uninfected samples; ONLY SIGNIFICANT INDICATOR TAXA
ita_ha_all_tbl = format_summary_table(ps_filt_all,
                                      ita_ha_all_significant,
                                      grp1_name = "HA Group 1",
                                      grp2_name = "HA Group 2",
                                      uninfected_name = "Uninfected")

# NA Groups including uninfected samples; ONLY SIGNIFICANT INDICATOR TAXA
ita_na_all_tbl = format_summary_table(ps_filt_all,
                                      ita_na_all_significant,
                                      grp1_name = "NA Group 1",
                                      grp2_name = "NA Group 2",
                                      uninfected_name = "Uninfected")

# HA Groups without uninfected samples; TOP 10 MOST SIGNIFICANT INDICATOR TAXA (none actually significant)
ita_ha_infected_tbl = format_summary_table(ps_filt_infected,
                                      ita_ha_infected,
                                      grp1_name = "HA Group 1",
                                      grp2_name = "HA Group 2",
                                      uninfected_name = NULL) %>%
  filter(Genus != "uncultured") %>%
  arrange("P Value", desc("Indicator Value")) %>%
  head(n = 10)

# NA Groups without uninfected samples; TOP 10 MOST SIGNIFICANT INDICATOR TAXA (none actually significant)
ita_na_infected_tbl = format_summary_table(ps_filt_infected,
                                      ita_na_infected,
                                      grp1_name = "NA Group 1",
                                      grp2_name = "NA Group 2",
                                      uninfected_name = NULL) %>%
  filter(Genus != "uncultured") %>%
  arrange("P Value", desc("Indicator Value")) %>%
  head(n = 10)
```

## 9. Save Tables

```{r}
# HA Groups including uninfected samples
save_ita_tbl(ita_ha_all_tbl, "ha_all")

# NA Groups including uninfected samples
save_ita_tbl(ita_na_all_tbl, "na_all")

# HA Groups without uninfected samples
save_ita_tbl(ita_ha_infected_tbl, "ha_infected")
  
# NA Groups without uninfected samples
save_ita_tbl(ita_na_infected_tbl, "na_infected")
```
