---
title: "indicator_taxa"
output: html_document
creator: "Alex Elbaz"
---

# Indicator Taxa Analysis

## 1. Load Libraries

```{r}
library(tidyverse)
library(phyloseq)
library(vegan)
library(ggpubr)
library(microbiome)
library(indicspecies)
library(gridExtra)
library(gtable)
library(grid)
```

## 2. Load Data

```{r}
# Read in phyloseq object
ps = readRDS('../processed_data/phyloseq_taxonomy.rds')
```

## 3. Set Seed & Prepare Data for Analyses

We deemed it most appropriate to agglomerate to the genus level for this analysis given the level of annotation available in the taxonomy data.

```{r}
# Set seed for repeatable analysis
set.seed(1234)

# Agglomerate taxa to genus level, convert to relative abundance, & apply abundance filter
ps_filt = tax_glom(ps, "Genus") %>%
  microbiome::transform("compositional") %>%
  filter_taxa(function(x) mean(x) >= 0.0001, TRUE)
```

## 4. Conduct Indicator Taxa Analysis (via indicpecies package)

An indicator value of \> 0.7 and p-value of \< 0.05 were considered as the thresholds for significance.

```{r}
# Conduct indicator taxa analysis by HA group & filter results for stat & p-value cutoffs
ita_ha = multipatt(t(otu_table(ps_filt)), 
                       cluster = sample_data(ps_filt)$ha_group,
                       control = how(nperm = 999))$sign %>%
  as.data.frame(.) %>%
  filter(stat > 0.7 & p.value < 0.05)

# Conduct indicator taxa analysis by NA group & filter results for stat & p-value cutoffs
ita_na = multipatt(t(otu_table(ps_filt)), 
                       cluster = sample_data(ps_filt)$na_group,
                       control = how(nperm = 999))$sign %>%
  as.data.frame(.) %>%
  filter(stat > 0.7 & p.value < 0.05)
```

## 5. Generate Indicator Taxa Summary Tables for HA & NA Analyses

The tables are organized such that each row corresponds to an indicator genus. The values of the columns "HA Group 1", "HA Group 2", "NA Group 1", "NA Group 2", and "Uninfected" represent whether the given genus is an indicator for that group (✔) or not (✘). Additionally, the tables are filtered to only include significant results (p \< 0.05 & stat \> 0.7).

### 5.1 HA Group Table

```{r}
# Get indicator genera names from indicator genera ASVs
indicator_genera_ha = data.frame(ps_filt@tax_table) %>%
  rownames_to_column() %>%
  filter(rowname %in% rownames(ita_ha)) %>%
  select(rowname, Genus)
indicator_genera_ha$Genus = indicator_genera_ha$Genus %>% str_replace("g__", "")

# Generate summary table
ita_summary_ha = ita_ha %>%
  rownames_to_column() %>%
  left_join(indicator_genera_ha)

# Format summary table for ease of interpretation
ita_summary_ha_formatted = ita_summary_ha %>%
  rename("HA Group 1" = "s.1", "HA Group 2" = "s.2", "Uninfected" = "s.neg") %>%
  select(-index, -rowname) %>%
  relocate(stat, .after = last_col()) %>%
  relocate(p.value, .before = stat) %>%
  relocate(Genus, .before = "HA Group 1") %>%
  mutate(stat = round(stat, 3)) %>%
  rename("Indicator Value" = stat, "P value" = p.value)

# Re-label indicator 0/1 markers with ✘/✔
for(col in colnames(ita_summary_ha_formatted[,2:4])) {
  ita_summary_ha_formatted[, col] = replace(ita_summary_ha_formatted[, col],
                                            which(1 == ita_summary_ha_formatted[, col]),
                                            "✔")
  ita_summary_ha_formatted[, col] = replace(ita_summary_ha_formatted[, col],
                                            which(0 == ita_summary_ha_formatted[, col]),
                                            "✘")
}

# Display formatted summary table
ita_summary_ha_formatted
```

### 5.2 NA Group Table

```{r}
# Get indicator genera names from indicator genera ASVs
indicator_genera_na = data.frame(ps_filt@tax_table) %>%
  rownames_to_column() %>%
  filter(rowname %in% rownames(ita_na)) %>%
  select(rowname, Genus)
indicator_genera_na$Genus = indicator_genera_na$Genus %>% str_replace("g__", "")

# Generate summary table
ita_summary_na = ita_na %>%
  rownames_to_column() %>%
  left_join(indicator_genera_na)

# Format summary table for ease of interpretation
ita_summary_na_formatted = ita_summary_na %>%
  rename("NA Group 1" = "s.1", "NA Group 2" = "s.2", "Uninfected" = "s.neg") %>%
  select(-index, -rowname) %>%
  relocate(stat, .after = last_col()) %>%
  relocate(p.value, .before = stat) %>%
  relocate(Genus, .before = "NA Group 1") %>%
  mutate(stat = round(stat, 3)) %>%
  rename("Indicator Value" = stat, "P value" = p.value)

# Re-label indicator 0/1 markers with ✘/✔
for(col in colnames(ita_summary_na_formatted[,2:4])) {
  ita_summary_na_formatted[, col] = replace(ita_summary_na_formatted[, col],
                                            which(1 == ita_summary_na_formatted[, col]),
                                            "✔")
  ita_summary_na_formatted[, col] = replace(ita_summary_na_formatted[, col],
                                            which(0 == ita_summary_na_formatted[, col]),
                                            "✘")
}

# Display formatted summary table
ita_summary_na_formatted
```

### 5.3 Save HA & NA Tables

```{r}
# HA Table
cairo_pdf("../visualizations/tables/ita_ha_summary.pdf")
ha_g = tableGrob(ita_summary_ha_formatted, theme = ttheme_minimal(base_size = 8), rows = NULL)
ha_g = gtable_add_grob(ha_g, grobs = segmentsGrob(x0 = unit(0, "npc"),
                                            y0 = unit(0, "npc"),
                                            x1 = unit(1, "npc"),
                                            y1 = unit(0, "npc")),
                    t = 1, l = 1, b = 1, r = ncol(ha_g))
grid.newpage()
grid.draw(ha_g)
dev.off()

# NA Table
cairo_pdf("../visualizations/tables/ita_na_summary.pdf")
na_g = tableGrob(ita_summary_na_formatted, theme = ttheme_minimal(base_size = 8), rows = NULL)
na_g = gtable_add_grob(na_g, grobs = segmentsGrob(x0 = unit(0, "npc"),
                                            y0 = unit(0, "npc"),
                                            x1 = unit(1, "npc"),
                                            y1 = unit(0, "npc")),
                    t = 1, l = 1, b = 1, r = ncol(na_g))
grid.newpage()
grid.draw(na_g)
dev.off()
```
