---
title: "random_forest"
author: "David"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries

```{r}
library(tidyverse)
library(dplyr)
library(phyloseq)
library(vegan)
library(ggpubr)
library(microbiome)
library(indicspecies)
library(writexl)
library(ANCOMBC)
library(broom)
library(randomForest)
library(caret)
library(ranger)
library(pROC)
library(boot)
library(MLmetrics)
```

## Loading Datafiles

To run our randomforest model we will use the same phyloseq object during earlier coding work. We will attempt to generate a randomforest predictive model using specific genera and their abundances based on previous results, including ...... . The MICB 305 reference script by Avril Metcalfe-Roach will be used to run a randomforest model.

```{r}
ps = readRDS('../processed_data/phyloseq_taxonomy.rds')|>
  tax_glom("Genus")

```

## Prepping our phyloseq object based on our identified genera of interest:

```{r}

genera_of_interest <- c("2d078b9d17ad33f59805e14ad64f8dfb", "a077b826f6681b86b0acda2a370692dc",
                        "527355921e32c01d361cc45757bf5d52")

ps_clr = ps |>
  microbiome::transform("clr")

ps_filt = prune_taxa(genera_of_interest, ps_clr)

df <- psmelt(ps_filt)

df_transformed <- df |>
  group_by("Genus") |>
  mutate(Abundance = scale(Abundance))|>
  ungroup()

df_pivot = df_transformed |>
  select(Sample, ha_group, Genus, Abundance) |>
  pivot_wider(names_from = Genus, values_from = Abundance)

df_noNA <- df_pivot |>
  na.omit()

df_final <- df_noNA|>
  select(-Sample)


```
## Running our RandomForest Model

```{r}

predictors <- df_final|>
  select(-ha_group)

df_final <- df_final |>
  mutate(ha_group = case_when(ha_group == "1" ~ "one", ha_group == "2" ~ "two", ha_group == "neg" ~ "neg"))

outcome = df_final|>
  pull(ha_group)|>
  factor(levels = c("one", "two", "neg"))

k = 10
set.seed(420)
folds = createFolds(outcome, k=k, list = TRUE)

str(folds)

tune_grid = expand.grid(mtry = c(2,3), splitrule = c(
  "gini"), min.node.size = c(100,150, 200)
)


source("reference_functions/randomforest_functions.R")

## Using Avril's Code (With some minor edits due to the neccesity for multiclass randomforest)

pd_model = run_rf(X= predictors, y = outcome, fold_list = folds, hyper = tune_grid, rngseed = 421)

names(pd_model)


```

## Visualizing our Results:

```{r}

roc_test = roc(pd_model$test_labels$true_labels,
               pd_model$test_labels$predicted_probabilities)
roc_train = roc(pd_model$train_labels$true_labels,
                pd_model$train_labels$predicted_probabilities)


ggplot() +
  geom_line(aes(x = 1 - roc_train$specificities, 
                y = roc_train$sensitivities), 
            color = "red",size=1) +
  geom_line(aes(x = 1 - roc_test$specificities,
                y = roc_test$sensitivities), 
            color = "black",size=1) +
  geom_abline(slope = 1, intercept = 0, color = "gray", linetype = "dashed",size=1) +
  labs(x = "False Positive Rate", y = "True Positive Rate") +
  annotate("text", x = 0.7, y = 0.2, 
           label = sprintf("Train (red): %.2f (%.2f-%.2f)\nTest (black): %.2f (%.2f-%.2f)",
           auc(roc_train), pd_model$auc_train_ci[1], pd_model$auc_train_ci[2],
           auc(roc_test), pd_model$auc_test_ci[1], pd_model$auc_test_ci[2]), 
           size = 6) +
  theme_minimal(base_size=18)

```