---
title: "random_forest"
author: "David"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries

```{r}
library(tidyverse)
library(dplyr)
library(phyloseq)
library(vegan)
library(ggpubr)
library(microbiome)
library(indicspecies)
library(writexl)
library(ANCOMBC)
library(broom)
library(randomForest)
library(caret)
library(ranger)
library(pROC)
library(boot)
library(MLmetrics)
```

## Loading Datafiles

To run our randomforest model we will use the same phyloseq object during earlier coding work. We will attempt to generate a randomforest predictive model using specific genera and their abundances based on previous results, including ...... . The MICB 305 reference script by Avril Metcalfe-Roach will be used to run a randomforest model.

```{r}
ps = readRDS('../processed_data/phyloseq_taxonomy.rds')|>
  tax_glom("Genus")

```

## Prepping our phyloseq object based on our identified genera of interest for HA:

```{r}

genera_of_interest <- c("dae4414c2d518a1a68ff10aeafb328fa", "ed2e115df9ed4a9cf8de15df330c428e",
                        "ac8a688dc06b4c9063048eea02c5d39f", "15efbfe72b43001ac4b74fa6a89cce83",
                        "2486ee9e808d35d25d6d5ebb4d3fdbba", "cda4e6f933bb3108ea3e92f9db411c00",
                        "24a60c6448e70d9198ad6ba93520958c", "e27680d4009f98f30248d823bc17fb8e",
                        "667fb5965fa7ccf7265f991dba0b5c5e", "fda8f480a378db2134a55c9dfb9b6b10",
                        "02bab2db0a7782cf48dec72cbac4bb20", "606cf1583385e5a58a195ff31b36a6d8")

ps_clr = ps |>
  microbiome::transform("clr")

ps_filt = prune_taxa(genera_of_interest, ps_clr)

df <- psmelt(ps_filt)

df_transformed <- df |>
  group_by("Genus") |>
  mutate(Abundance = scale(Abundance))|>
  ungroup()

df_pivot = df_transformed |>
  select(Sample, ha_group, Genus, Abundance) |>
  pivot_wider(names_from = Genus, values_from = Abundance)

## We will also filter for only group 1 and group 2, as we are interested in seeing if we can predict group 1 vs group 2 rather than group 1 vs control or group 2 vs control.

df_noNA <- df_pivot |>
  na.omit()|>
  filter(ha_group != "neg")|>
  mutate(ha_group = case_when(ha_group == 1 ~ "one", ha_group == 2 ~ "two"))|>
  mutate(ha_group = as.factor(ha_group))

df_final <- df_noNA|>
  select(-Sample)


```
## Prepping our phyloseq object based on our identified genera of interest for NA:

```{r}

genera_of_interest_for_na <- c("fda8f480a378db2134a55c9dfb9b6b10", "ed2e115df9ed4a9cf8de15df330c428e",
                        "e27680d4009f98f30248d823bc17fb8e", "dd90232e0889025ff448018185547d6c",
                        "dae4414c2d518a1a68ff10aeafb328fa", "cda4e6f933bb3108ea3e92f9db411c00",
                        "ac8a688dc06b4c9063048eea02c5d39f", "24a60c6448e70d9198ad6ba93520958c",
                        "15efbfe72b43001ac4b74fa6a89cce83", "02bab2db0a7782cf48dec72cbac4bb20",
                        "667fb5965fa7ccf7265f991dba0b5c5e", "2486ee9e808d35d25d6d5ebb4d3fdbba",
                        "606cf1583385e5a58a195ff31b36a6d8")

ps_clr = ps |>
  microbiome::transform("clr")

ps_filt_na = prune_taxa(genera_of_interest_for_na, ps_clr)

df_na <- psmelt(ps_filt_na)

df_transformed_na <- df_na |>
  group_by("Genus") |>
  mutate(Abundance = scale(Abundance))|>
  ungroup()

df_pivot_na = df_transformed_na |>
  select(Sample, na_group, Genus, Abundance) |>
  pivot_wider(names_from = Genus, values_from = Abundance)

## We will also filter for only group 1 and group 2, as we are interested in seeing if we can predict group 1 vs group 2 rather than group 1 vs control or group 2 vs control.

df_noHA <- df_pivot_na |>
  na.omit()|>
  filter(na_group != "neg")|>
  mutate(na_group = case_when(na_group == 1 ~ "one", na_group == 2 ~ "two"))|>
  mutate(na_group = as.factor(na_group))

df_final_na <- df_noHA|>
  select(-Sample)
```

## Running our RandomForest Model for HA and NA

```{r}

predictors_na <- df_final_na|>
  select(-na_group)

outcome_na = df_final_na|>
  pull(na_group)

k = 10
set.seed(420)
folds = createFolds(outcome_na, k=k, list = TRUE)

str(folds)

tune_grid = expand.grid(mtry = c(3,9, 10), splitrule = c(
  "gini"), min.node.size = c(100,150, 200)
)


source("reference_functions/randomforest_functions.R")

## Using Avril's Code from MICB 305 repository

pd_model = run_rf(X= predictors_na, y = outcome_na, fold_list = folds, hyper = tune_grid, rngseed = 421)

names(pd_model)

## running the code for HA, simply by replacing the variable names.

predictors <- df_final|>
  select(-ha_group)

outcome = df_final|>
  pull(ha_group)

k = 10
set.seed(420)
folds = createFolds(outcome, k=k, list = TRUE)

str(folds)

tune_grid = expand.grid(mtry = c(3,10, 12), splitrule = c(
  "gini"), min.node.size = c(100,150, 200)
)


source("reference_functions/randomforest_functions.R")

## Using Avril's Code (With some minor edits due to the neccesity for multiclass randomforest)

pd_model = run_rf(X= predictors, y = outcome, fold_list = folds, hyper = tune_grid, rngseed = 421)

names(pd_model)

```

## Visualizing our Results for both HA and NA:

```{r}

roc_test = roc(pd_model$test_labels$true_labels,
               pd_model$test_labels$predicted_probabilities)
roc_train = roc(pd_model$train_labels$true_labels,
                pd_model$train_labels$predicted_probabilities)

roc_data = data.frame(Dataset = 'df',
     Training_AUC = round(pd_model$auc_train, 2),
     Training_AUC_CI = paste0(round(pd_model$auc_train_ci[1], 2), "-", 
                              round(pd_model$auc_train_ci[2], 2)),
     Testing_AUC = round(pd_model$auc_test, 2),
     Testing_AUC_CI = paste0(round(pd_model$auc_test_ci[1], 2), "-", 
                             round(pd_model$auc_test_ci[2], 2)))


na_auc <- ggplot() +
  geom_line(aes(x = 1 - roc_train$specificities, 
                y = roc_train$sensitivities), 
            color = "red",size=1) +
  geom_line(aes(x = 1 - roc_test$specificities,
                y = roc_test$sensitivities), 
            color = "black",size=1) +
  geom_abline(slope = 1, intercept = 0, color = "gray", linetype = "dashed",size=1) +
  labs(x = "False Positive Rate", y = "True Positive Rate") +
  annotate("text", x = 0.7, y = 0.2, 
           label = sprintf("Train (red): %.2f (%.2f-%.2f)\nTest (black): %.2f (%.2f-%.2f)",
           auc(roc_train), pd_model$auc_train_ci[1], pd_model$auc_train_ci[2],
           auc(roc_test), pd_model$auc_test_ci[1], pd_model$auc_test_ci[2]), 
           size = 6) +
  theme_minimal(base_size=24)

na_auc

```
## Generating an importance graph for NA randomforest:

``` {r}

na_rf_importance_graph <- pd_model$importance %>% 
  mutate(Feature = factor(.$Feature,levels = .$Feature)) %>% 
  ggplot(aes(Feature,MeanDecreaseGini,fill=MeanDecreaseGini)) +
  geom_col() +
  theme_classic(base_size=18) +
  theme(axis.text.x = element_text(angle=45, vjust = 1, hjust=1)) +
  ylab('Importance (Gini)') + xlab(NULL)

na_rf_importance_graph

colnames(df_final_na)

temp = ps_filt_na %>% 
  subset_taxa(Genus==' g__Pedobacter') %>% 
  psmelt()

temp <- temp |>
  filter(na_group != "neg")

wilcox.test(Abundance ~ na_group, data = temp)

pedobacter_graph = temp %>% ggplot(aes(na_group,Abundance)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height=0, width = 0.25, alpha = .2) +
  theme_classic(base_size=16) +
  geom_signif(comparisons = list(c("1", "2")), map_signif_level = TRUE, textsize = 6)+
  labs(x = "NA group", y = "CLR-Transformed Abundance") +
  theme(axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        axis.text = element_text(size = 16))+
  ylim(-3, 15)

pedobacter_graph


```

## Saving our graphs
```{r}

ggsave("../visualizations/figures/ha_randomforest.png", plot = ha_auc, width = 18, height = 10)
ggsave("../visualizations/figures/na_randomforest_auc.png", plot = na_auc, width = 18, height = 10)
ggsave("../visualizations/figures/na_randomforest_importance.png", plot = na_rf_importance_graph, width = 18, height = 10)
ggsave("../visualizations/figures/na_randomforest_pedobacter.png", plot = pedobacter_graph, width = 18, height = 10)



```