---
title: "random_forest"
author: "David"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries

```{r}
library(tidyverse)
library(mixOmics)
library(dplyr)
library(phyloseq)
library(vegan)
library(ggpubr)
library(microbiome)
library(indicspecies)
library(writexl)
library(ANCOMBC)
library(broom)
library(randomForest)
library(caret)
library(ranger)
library(pROC)
library(boot)
library(MLmetrics)
```

## Loading Datafiles

To run our randomforest model we will use the same phyloseq object during earlier coding work. We will attempt to generate a randomforest predictive model using VSURF predicted taxa of importance. The MICB 305 reference script by Avril Metcalfe-Roach will be used to run a randomforest model. All HA related scripts that would interfere with NA related scripts have been commented out. Any person who wishes to run our analyses on HA data should comment out NA related scripts while uncommenting HA related scripts.

```{r}
ps = readRDS('../processed_data/phyloseq_taxonomy.rds')|>
  tax_glom("Genus")

```

## Wrangling our PS object for HA and NA RF model

```{r}

select <- dplyr::select

set.seed(420)

ps_clr = ps |>
  microbiome::transform("clr")|>
  filter_taxa(function(x) mean(x) >= 0.0001, TRUE)

vsurf_df <- psmelt(ps_clr)

vsurf_df_transformed <- vsurf_df |>
  group_by(Genus)|>
  mutate(Abundance = scale(Abundance))|>
  ungroup()|>
  mutate(Abundance = Abundance[, 1])

vsurf_df_transformed <- vsurf_df_transformed |>
    filter(Genus != " g__uncultured")

vsurf_df_transformed$Genus = vsurf_df_transformed$Genus %>% str_replace("g__", "")


 vsurf_df_pivot_ha <- vsurf_df_transformed |>
   select(Sample, ha_group, Genus, Abundance) |>
   pivot_wider(names_from = Genus, values_from = Abundance)|>
   filter(ha_group != "neg")|>
   select(-Sample)

vsurf_df_pivot_na = vsurf_df_transformed |>
  select(Sample, na_group, Genus, Abundance) |>
  pivot_wider(names_from = Genus, values_from = Abundance)|>
  filter(na_group != "neg")|>
  select(-Sample)



```


## Prepping our HA df for RF:

```{r}

# ha_prepped_rf_df <- vsurf_df_pivot_ha |>
#   na.omit()|>
#   filter(ha_group != "neg")|>
#   mutate(ha_group = case_when(ha_group == 1 ~ "one", ha_group == 2 ~ "two"))|>
#   mutate(ha_group = as.factor(ha_group))

```
## Prepping our NA df for RF:

```{r}

na_prepped_rf_df <- vsurf_df_pivot_na |>
  na.omit()|>
  filter(na_group != "neg")|>
  mutate(na_group = case_when(na_group == 1 ~ "one", na_group == 2 ~ "two"))|>
  mutate(na_group = as.factor(na_group))
# # 
```
# 
# ## Running our RandomForest Model for HA and NA
# 
```{r}
# 
# ## running code for NA/HA RF (THROUGH MANUAL REPLACEMENT)
# 

## HA RF

# predictors_ha <- ha_prepped_rf_df|>
#   select(-ha_group)
# 
# outcome_ha = ha_prepped_rf_df|>
#   pull(ha_group)
# 
# k = 10
# folds = createFolds(outcome_ha, k=k, list = TRUE)
# 
# tune_grid = expand.grid(mtry = c(7, 15, 20), splitrule = c(
# "gini"), min.node.size = c(10,25,50))
# 
# source("reference_functions/randomforest_functions.R")
# 
# ## Using Avril's Code from MICB 305 repository
# 
# pd_model = run_rf(X= predictors_ha, y = outcome_ha, fold_list = folds, hyper = tune_grid, rngseed = 421)

## NA RF

predictors_na <- na_prepped_rf_df|>
  select(-na_group)

outcome_na = na_prepped_rf_df|>
  pull(na_group)

k = 10
folds = createFolds(outcome_na, k=k, list = TRUE)

tune_grid = expand.grid(mtry = c(7, 15, 20), splitrule = c(
  "gini"), min.node.size = c(10,25,50))

source("reference_functions/randomforest_functions.R")

## Using Avril's Code from MICB 305 repository

pd_model = run_rf(X= predictors_na, y = outcome_na, fold_list = folds, hyper = tune_grid, rngseed = 421)

```

## Visualizing our Results for both HA and NA:

```{r}

roc_test = roc(pd_model$test_labels$true_labels,
               pd_model$test_labels$predicted_probabilities)
roc_train = roc(pd_model$train_labels$true_labels,
                pd_model$train_labels$predicted_probabilities)

roc_data = data.frame(Dataset = 'df',
     Training_AUC = round(pd_model$auc_train, 2),
     Training_AUC_CI = paste0(round(pd_model$auc_train_ci[1], 2), "-",
                              round(pd_model$auc_train_ci[2], 2)),
     Testing_AUC = round(pd_model$auc_test, 2),
     Testing_AUC_CI = paste0(round(pd_model$auc_test_ci[1], 2), "-",
                             round(pd_model$auc_test_ci[2], 2)))


#### HA SPECIFIC


# ha_auc <- ggplot() +
#   geom_line(aes(x = 1 - roc_train$specificities,
#                 y = roc_train$sensitivities),
#             color = "red",size=1) +
#   geom_line(aes(x = 1 - roc_test$specificities,
#                 y = roc_test$sensitivities),
#             color = "black",size=1) +
#   geom_abline(slope = 1, intercept = 0, color = "gray", linetype = "dashed",size=1) +
#   labs(x = "", y = "") +
#   annotate("text", x = 0.7, y = 0.2,
#            label = sprintf("Train (red): %.2f (%.2f-%.2f)\nTest (black): %.2f (%.2f-%.2f)",
#            auc(roc_train), pd_model$auc_train_ci[1], pd_model$auc_train_ci[2],
#            auc(roc_test), pd_model$auc_test_ci[1], pd_model$auc_test_ci[2]),
#            size = 12) +
#   theme_minimal(base_size=24)
# 
# ha_auc


#### NA SPECIFIC

na_auc <- ggplot() +
  geom_line(aes(x = 1 - roc_train$specificities,
                y = roc_train$sensitivities),
            color = "red",size=1) +
  geom_line(aes(x = 1 - roc_test$specificities,
                y = roc_test$sensitivities),
            color = "black",size=1) +
  geom_abline(slope = 1, intercept = 0, color = "gray", linetype = "dashed",size=1) +
  labs(x = "", y = "") +
  annotate("text", x = 0.7, y = 0.2,
           label = sprintf("Train (red): %.2f (%.2f-%.2f)\nTest (black): %.2f (%.2f-%.2f)",
           auc(roc_train), pd_model$auc_train_ci[1], pd_model$auc_train_ci[2],
           auc(roc_test), pd_model$auc_test_ci[1], pd_model$auc_test_ci[2]),
           size = 12) +
  labs(x = "False Positive Rate", y = "True Positive Rate")+
  theme_classic(base_size=24)

na_auc


```
## Generating an importance graph for NA/HA randomforest:

``` {r}

## NA graphs

pd_model$importance <- pd_model$importance |>
   filter(MeanDecreaseGini >= .6)

na_rf_importance_graph <- pd_model$importance %>%
  mutate(Feature = factor(.$Feature,levels = .$Feature)) %>%
  ggplot(aes(Feature,MeanDecreaseGini,fill=MeanDecreaseGini)) +
  geom_col() +
  theme_classic(base_size=28) +
  theme(axis.text.x = element_text(angle=70, vjust = 1, hjust=1)) +
  ylab('Importance (Gini)') + xlab(NULL)

na_rf_importance_graph


## HA Graph

# pd_model$importance <- pd_model$importance |>
#     filter(MeanDecreaseGini >= .6)
# 
# ha_rf_importance_graph <- pd_model$importance %>%
#   mutate(Feature = factor(.$Feature,levels = .$Feature)) %>%
#   ggplot(aes(Feature,MeanDecreaseGini,fill=MeanDecreaseGini)) +
#   geom_col() +
#   theme_classic(base_size=28) +
#   theme(axis.text.x = element_text(angle=90, vjust = 1, hjust=1)) +
#   ylab('Importance (Gini)') + xlab(NULL)
# 
# ha_rf_importance_graph
# 


```

## Saving our graphs
```{r}
# 
# ggsave("../visualizations/figures/ha_randomforest_auc.png", plot = ha_auc, width = 18, height = 10)
# ggsave("../visualizations/figures/ha_randomforest_importance.png", plot = ha_rf_importance_graph, width = 18, height = 10)


ggsave("../visualizations/figures/na_randomforest_auc.png", plot = na_auc, width = 18, height = 10)
ggsave("../visualizations/figures/na_randomforest_importance.png", plot = na_rf_importance_graph, width = 18, height = 10)





```
