---
title: "random_forest"
output: html_document
author: "David Medvedev"
---

## Load Libraries and Reference Functions

```{r}
library(tidyverse)
library(mixOmics)
library(dplyr)
library(phyloseq)
library(vegan)
library(ggpubr)
library(microbiome)
library(indicspecies)
library(writexl)
library(ANCOMBC)
library(broom)
library(randomForest)
library(caret)
library(ranger)
library(pROC)
library(boot)
library(MLmetrics)

## Using Avril's Code from MICB 305 repository
source("reference_functions/randomforest_functions.R")
```

## Load Data

To build our RandomForest models we will use the previously generated phyloseq object alongside the reference R script created for MICB 305 by Dr. Avril Metcalfe-Roach.

```{r}
ps = readRDS('../processed_data/phyloseq_taxonomy.rds')|> tax_glom("Genus")
```

## Wrangle Phyloseq Object for HA and NA RandomForest Models

```{r}
select <- dplyr::select
set.seed(420)


ps_clr = ps |>
  microbiome::transform("clr")|>
  filter_taxa(function(x) mean(x) >= 0.0001, TRUE)


rf_df_transformed <- psmelt(ps_clr) |>
  group_by(Genus)|>
  mutate(Abundance = scale(Abundance))|>
  ungroup()|>
  mutate(Abundance = Abundance[, 1]) |>
  filter(Genus != " g__uncultured")

rf_df_transformed$Genus = rf_df_transformed$Genus %>% str_replace("g__", "")


rf_df_ha <- rf_df_transformed |>
   select(Sample, ha_group, Genus, Abundance) |>
   pivot_wider(names_from = Genus, values_from = Abundance)|>
   filter(ha_group != "neg")|>
   select(-Sample) |>
   na.omit()|>
   mutate(ha_group = case_when(ha_group == 1 ~ "one", ha_group == 2 ~ "two"))|>
   mutate(ha_group = as.factor(ha_group))

rf_df_na = rf_df_transformed |>
  select(Sample, na_group, Genus, Abundance) |>
  pivot_wider(names_from = Genus, values_from = Abundance)|>
  filter(na_group != "neg")|>
  select(-Sample) |>
  na.omit()|>
  mutate(na_group = case_when(na_group == 1 ~ "one", na_group == 2 ~ "two"))|>
  mutate(na_group = as.factor(na_group))
```

## Generate RandomForest Models for HA and NA

```{r}
# Create predictors and outcomes
predictors_ha <- rf_df_ha |>
  select(-ha_group)
outcome_ha = rf_df_ha |>
  pull(ha_group)

predictors_na <- rf_df_na|>
  select(-na_group)
outcome_na = rf_df_na|>
  pull(na_group)

# Set up cross-validation and hyperparameter optimization
k = 10
ha_folds = createFolds(outcome_ha, k=k, list = TRUE)
na_folds = createFolds(outcome_na, k=k, list = TRUE)

tune_grid = expand.grid(mtry = c(7, 15, 20), splitrule = c(
"gini"), min.node.size = c(10,25,50))


# Build RandomForest models
ha_model = run_rf(X=predictors_ha, y=outcome_ha, fold_list=ha_folds, hyper=tune_grid, rngseed=421)
na_model = run_rf(X=predictors_na, y=outcome_na, fold_list=na_folds, hyper=tune_grid, rngseed=421)
```

## Visualize Results

### HA

```{r}
roc_test_ha = roc(ha_model$test_labels$true_labels,
               ha_model$test_labels$predicted_probabilities)
roc_train_ha = roc(ha_model$train_labels$true_labels,
                ha_model$train_labels$predicted_probabilities)

roc_data_ha = data.frame(Dataset = 'df',
     Training_AUC = round(ha_model$auc_train, 2),
     Training_AUC_CI = paste0(round(ha_model$auc_train_ci[1], 2), "-",
                              round(ha_model$auc_train_ci[2], 2)),
     Testing_AUC = round(ha_model$auc_test, 2),
     Testing_AUC_CI = paste0(round(ha_model$auc_test_ci[1], 2), "-",
                             round(ha_model$auc_test_ci[2], 2)))


ha_auc <- ggplot() +
  geom_line(aes(x = 1 - roc_train_ha$specificities,
                y = roc_train_ha$sensitivities),
            color = "red",size=1) +
  geom_line(aes(x = 1 - roc_test_ha$specificities,
                y = roc_test_ha$sensitivities),
            color = "black",size=1) +
  geom_abline(slope = 1, intercept = 0, color = "gray", linetype = "dashed",size=1) +
  annotate("text", x = 0.7, y = 0.2,
           label = sprintf("Train (red): %.2f (%.2f-%.2f)\nTest (black): %.2f (%.2f-%.2f)",
           auc(roc_train_ha), ha_model$auc_train_ci[1], ha_model$auc_train_ci[2],
           auc(roc_test_ha), ha_model$auc_test_ci[1], ha_model$auc_test_ci[2]),
           size = 12) +
  labs(x = "False Positive Rate", y = "True Positive Rate")+
  theme_classic(base_size=24)

ha_auc
```

### NA

```{r}
roc_test_na = roc(na_model$test_labels$true_labels,
               na_model$test_labels$predicted_probabilities)
roc_train_na = roc(na_model$train_labels$true_labels,
                na_model$train_labels$predicted_probabilities)

roc_data_na = data.frame(Dataset = 'df',
     Training_AUC = round(na_model$auc_train, 2),
     Training_AUC_CI = paste0(round(na_model$auc_train_ci[1], 2), "-",
                              round(na_model$auc_train_ci[2], 2)),
     Testing_AUC = round(na_model$auc_test, 2),
     Testing_AUC_CI = paste0(round(na_model$auc_test_ci[1], 2), "-",
                             round(na_model$auc_test_ci[2], 2)))


na_auc <- ggplot() +
  geom_line(aes(x = 1 - roc_train_na$specificities,
                y = roc_train_na$sensitivities),
            color = "red",size=1) +
  geom_line(aes(x = 1 - roc_test_na$specificities,
                y = roc_test_na$sensitivities),
            color = "black",size=1) +
  geom_abline(slope = 1, intercept = 0, color = "gray", linetype = "dashed",size=1) +
  annotate("text", x = 0.7, y = 0.2,
           label = sprintf("Train (red): %.2f (%.2f-%.2f)\nTest (black): %.2f (%.2f-%.2f)",
           auc(roc_train_na), na_model$auc_train_ci[1], na_model$auc_train_ci[2],
           auc(roc_test_na), na_model$auc_test_ci[1], na_model$auc_test_ci[2]),
           size = 12) +
  labs(x = "False Positive Rate", y = "True Positive Rate") +
  theme_classic(base_size=24)

na_auc
```

## Generate Importance Plots for HA and NA Models

### HA

```{r}
ha_model$importance <- ha_model$importance |> filter(MeanDecreaseGini >= .6)

ha_rf_importance_graph <- ha_model$importance %>%
  mutate(Feature = factor(.$Feature,levels = .$Feature)) %>%
  ggplot(aes(Feature,MeanDecreaseGini,fill=MeanDecreaseGini)) +
  geom_col() +
  theme_classic(base_size=28) +
  theme(axis.text.x = element_text(angle=70, vjust = 1, hjust=1)) +
  ylab('Importance (Gini)') + xlab(NULL)

ha_rf_importance_graph
```

### NA

```{r}
na_model$importance <- na_model$importance |>
   filter(MeanDecreaseGini >= 0.6266017)

na_rf_importance_graph <- na_model$importance %>%
  mutate(Feature = factor(.$Feature,levels = .$Feature)) %>%
  ggplot(aes(Feature,MeanDecreaseGini,fill=MeanDecreaseGini)) +
  geom_col() +
  theme_classic(base_size=28) +
  theme(axis.text.x = element_text(angle=70, vjust = 1, hjust=1)) +
  ylab('Importance (Gini)') + xlab(NULL)

na_rf_importance_graph
```

## Save Plots

```{r}
ggsave("../visualizations/figures/random_forest/ha_randomforest_auc.png", plot = ha_auc, width = 18, height = 10)
ggsave("../visualizations/figures/random_forest/ha_randomforest_importance.png", plot = ha_rf_importance_graph, width = 18, height = 10)

ggsave("../visualizations/figures/random_forest/na_randomforest_auc.png", plot = na_auc, width = 18, height = 10)
ggsave("../visualizations/figures/random_forest/na_randomforest_importance.png", plot = na_rf_importance_graph, width = 18, height = 10)
```
